---
title: "Single-ordering playground"
output:
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                     fig.width = 11, fig.height = 7)
# Prefer the installed package. Set MPCURVER_LIB to use a particular library.
if (nzchar(Sys.getenv("MPCURVER_LIB")))
  .libPaths(c(Sys.getenv("MPCURVER_LIB"), .libPaths()))
library(MPCurver)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

## Intro

Change `config` or replace `simulate_signal()`. The remaining chunks use the
same interface: a sample-by-feature signal matrix evaluated at `t`.

```{r configuration}
config <- list(
  seed = 20260932L, n = 300L, p = 12L, snr = 4,
  generator = "matern",       # "matern" or "fourier"
  length_scale = 0.2,         # Matern 5/2
  frequencies = 2:6, decay = 1.5,  # Fourier
  reference = "V1",           # Figure 2 horizontal axis
  K = 50L, position_prior = "fixed", # "fixed" = uniform; or "adaptive"
  max_iter = 2000L, tol = 1e-6
)
```

```{r simulation-module}
simulate_signal <- function(t, p, config) {
  if (config$generator == "matern") {
    distance <- abs(outer(t, t, "-"))
    z <- sqrt(5) * distance / config$length_scale
    covariance <- (1 + z + z^2 / 3) * exp(-z)
    L <- t(chol(covariance + diag(1e-10, length(t))))
    return(L %*% matrix(rnorm(length(t) * p), length(t), p))
  }
  if (config$generator == "fourier") {
    h <- config$frequencies
    angles <- outer(t, pi * h)
    a <- matrix(rnorm(length(h) * p), length(h), p)
    b <- matrix(rnorm(length(h) * p), length(h), p)
    return(sin(angles) %*% (h^(-config$decay) * a) +
           cos(angles) %*% (h^(-config$decay) * b))
  }
  stop("Unknown generator: ", config$generator)
}

simulate_data <- function(config) {
  stopifnot(config$n > config$K, config$p >= 1, config$snr > 0)
  set.seed(config$seed)
  t <- runif(config$n)
  signal <- simulate_signal(t, config$p, config)
  stopifnot(identical(dim(signal), c(config$n, config$p)),
            all(is.finite(signal)), all(apply(signal, 2, sd) > 0))
  signal <- scale(signal)  # unit signal variance per column
  # Separate noise seed keeps noise fixed when changing the generator.
  set.seed(config$seed + 1L)
  X <- signal + matrix(rnorm(length(signal)), nrow(signal)) / sqrt(config$snr)
  colnames(X) <- colnames(signal) <- paste0("V", seq_len(config$p))
  list(t = t, signal = signal, X = X)
}
data <- simulate_data(config)
```

```{r plotting-module}
scale01 <- function(x) {
  stopifnot(diff(range(x)) > 0)
  (x - min(x)) / diff(range(x))
}
recovery <- function(x, truth) abs(cor(x, truth, method = "spearman"))

feature_frame <- function(values, position, truth) {
  do.call(rbind, lapply(seq_len(ncol(values)), function(j)
    data.frame(position = position, value = values[, j], truth = truth,
               feature = factor(colnames(values)[j], levels = colnames(values)))))
}

plot_features <- function(data, position, xlabel, signal_position = NULL) {
  observed <- feature_frame(data$X, position, data$t)
  if (is.null(signal_position)) {
    ggplot(observed, aes(position, value, color = truth)) +
      geom_point(size = 0.8, alpha = 0.65) +
      scale_color_viridis_c(name = "True t", limits = c(0, 1)) +
      facet_wrap(~feature, ncol = 4) + labs(x = xlabel, y = "Observed value")
  } else {
    idx <- order(data$t)
    paths <- feature_frame(data$signal[idx, , drop = FALSE],
                           signal_position[idx], data$t[idx])
    ggplot(observed, aes(position, value)) +
      geom_point(color = "gray55", size = 0.6, alpha = 0.3) +
      geom_path(data = paths, color = "#0072B2", linewidth = 0.8) +
      facet_wrap(~feature, ncol = 4) + labs(x = xlabel, y = "Feature value")
  }
}
```

```{r figure-1, fig.cap="Figure 1. Features versus true t. Blue paths join noiseless values in true-position order; gray points are observations."}
plot_features(data, data$t, "True latent position", signal_position = data$t)
```

```{r figure-2, fig.cap="Figure 2. Features versus the reference column. Blue paths are noiseless pairs; gray points contain noise on both axes."}
j <- match(config$reference, colnames(data$X))
stopifnot(!is.na(j))
plot_features(data, data$X[, j], config$reference,
              signal_position = data$signal[, j])
```

## PCA

PC1 uses centered, unscaled observed columns. Its sign is aligned with truth
only for display; positions retain score spacing.

```{r pca}
pca <- prcomp(data$X, center = TRUE, scale. = FALSE)
pc_raw <- pca$x[, 1]
direction <- if (cor(pc_raw, data$t, method = "spearman") < 0) -1 else 1
pc_position <- scale01(direction * pc_raw)
```

```{r figure-3, fig.cap="Figure 3. Observations versus scaled PC1 scores; colors indicate true t."}
plot_features(data, pc_position, "PC1 scores scaled to [0,1]")
```

```{r figure-4, fig.height=5, fig.cap="Figure 4. PCA positions versus true t. Dashed line: identity."}
ggplot(data.frame(truth = data$t, estimate = pc_position), aes(truth, estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray70") +
  geom_point(alpha = 0.65, size = 1.2) + coord_equal() +
  labs(x = "True latent position", y = "Scaled PC1 position",
       title = sprintf("PCA: |Spearman rho| = %.3f", recovery(pc_position, data$t)))
```

## MPCurve

Single-ordering CAVI, initialized from PCA using quantile bins. Fixed prior
means probability `1/K` throughout fitting. Uses MPCurver 0.3.4 internal helpers.

```{r fitting-module}
fit_from_positions <- function(X, position, config) {
  init <- MPCurver:::.cavi_exact_k_init_from_ordering(
    X = X, ordering_vec = position, K = config$K,
    discretization = "quantile", strict_K = TRUE)
  gamma0 <- MPCurver:::.cavi_hard_gamma_from_cluster_rank(
    init$cluster_rank, nrow(X), config$K)
  MPCurver:::cavi(
    X = X, K = config$K, responsibilities_init = gamma0,
    position_prior_init = rep(1 / config$K, config$K),
    position_prior = config$position_prior, sigma2_init = init$sigma2,
    lambda_init = rep(1, ncol(X)), rw_q = 2L, ridge = 0,
    discretization = "quantile", max_iter = config$max_iter,
    tol = config$tol, convergence = "relative", verbose = FALSE)
}
fit <- fit_from_positions(data$X, pc_raw, config)
mp_position <- as.numeric(fit$gamma %*% seq(0, 1, length.out = config$K))
if (direction < 0) mp_position <- 1 - mp_position
initial_elbo <- head(fit$elbo_trace, 1)
final_elbo <- tail(fit$elbo_trace, 1)
# Keep convergence status visible even when the iteration budget is exhausted.
data.frame(version = as.character(packageVersion("MPCurver")),
           iterations = fit$iter, converged = fit$converged,
           initial_ELBO = initial_elbo, final_ELBO = final_elbo)
```

```{r figure-5, fig.cap="Figure 5. Observations versus MPCurve posterior mean positions; colors indicate true t."}
plot_features(data, mp_position, "MPCurve posterior mean position")
```

```{r figure-6, fig.height=5.8, fig.cap="Figure 6. Before/after comparison against true t, with the same orientation. Initial ELBO is evaluated after PCA quantile discretization and variational initialization, before CAVI updates."}
labels <- c(
  sprintf("PCA initialization\nrho = %.3f | ELBO = %.2f",
          recovery(pc_position, data$t), initial_elbo),
  sprintf("MPCurve: %d updates, converged = %s\nrho = %.3f | ELBO = %.2f",
          fit$iter, fit$converged, recovery(mp_position, data$t), final_elbo))
comparison <- data.frame(
  truth = rep(data$t, 2), estimate = c(pc_position, mp_position),
  stage = factor(rep(labels, each = length(data$t)), levels = labels))
ggplot(comparison, aes(truth, estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray70") +
  geom_point(alpha = 0.65, size = 1.2) + facet_wrap(~stage, nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "True latent position", y = "Estimated latent position")
```

```{r provenance, include=FALSE}
# Available in the knitted session for saving/reusing results.
experiment <- list(config = config, data = data, pca = pca, fit = fit,
                   pc_position = pc_position, mp_position = mp_position,
                   session = sessionInfo())
```
