---
title: "Explore the swiss roll"
author: "Ziang Zhang"
date: "2025-12-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
## Introduction

In this example, we will consider how to learn a 1D manifold (swiss-roll/spiral) embedded in 2D with noise, using various variants of principal curve fitting and our SmoothEM algorithm.

Simulate a 1D swiss-roll (spiral) embedded in 2D with noise.

```{r}
simulate_swiss_roll_1d_2d <- function(n = 800,
                                     t_range = c(1.5 * pi, 6 * pi),
                                     sigma = 0.15,     # can be scalar or length-2
                                     seed = 1) {
  stopifnot(length(t_range) == 2)
  stopifnot(length(sigma) == 1 || length(sigma) == 2)

  set.seed(seed)

  # 1D parameter
  t <- runif(n, min = t_range[1], max = t_range[2])

  # true 2D "swiss roll" (spiral)
  x_true <- t * cos(t)
  y_true <- t * sin(t)

  # noise
  if (length(sigma) == 1) sigma <- rep(sigma, 2)
  x_obs <- x_true + rnorm(n, 0, sigma[1])
  y_obs <- y_true + rnorm(n, 0, sigma[2])

  list(
    t = t,
    truth = data.frame(x = x_true, y = y_true),
    obs   = data.frame(x = x_obs,  y = y_obs)
  )
}

# Example
sim <- simulate_swiss_roll_1d_2d(n = 1500, sigma = 0.2, seed = 123)

# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y", main = "Noisy 2D swiss-roll (spiral)")
# Colors mapped from t
pal <- colorRampPalette(c("navy", "cyan", "yellow", "red"))(256)
t_scaled <- (sim$t - min(sim$t)) / (max(sim$t) - min(sim$t))  # in [0,1]
col_t <- pal[pmax(1, pmin(256, 1 + floor(255 * t_scaled)))]
# Add truth, colored by t
points(sim$truth$x, sim$truth$y, pch = 16, cex = 0.25, col = col_t)
```


## Principal Curve Fitting

The simpliest principal curve is just using the initial PCA line (zero iterations).

```{r}
library(princurve)
# Initial fit with zero iterations (just the initial PCA line)
pc_fit <- principal_curve(as.matrix(sim$obs), maxit = 0)
# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
# Add principal curve
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
# Add arrow
arrows(pc_fit$s[1, 1], pc_fit$s[1, 2], pc_fit$s[2, 1], pc_fit$s[2, 2])
```

Apply principal curve fitting to recover the 1D structure with the default settings.

```{r}
set.seed(123)
pc_fit <- principal_curve(as.matrix(sim$obs))

# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
# Add principal curve (sorted by lambda!)
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
```

Try to increase fitting accuracy by lowering the convergence threshold.

```{r}
set.seed(123)
pc_fit <- principal_curve(as.matrix(sim$obs), thresh = 0.0001, 
                          stretch = 0, maxit = 100)

plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
```


Try a different smoother: lowess.

```{r}
set.seed(123)
pc_fit_lowess <- principal_curve(as.matrix(sim$obs), smoother = "lowess", 
                                 thresh = 0.0001,
                                 maxit = 1000)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit (lowess) to Noisy Swiss-Roll",
     asp = 1)
ord <- order(pc_fit_lowess$lambda)
lines(pc_fit_lowess$s[ord, 1], pc_fit_lowess$s[ord, 2], col = "black", lwd = 2)
```


## SmoothEM

None of the above principal curve fittings are able to recover the 1D swiss-roll structure well, even when we switch to different smoothers.

Most likely it is due to the improper initialization using PCA line.

Now let's try various way to initialize the SmoothEM algorithm to recover the 1D structure.

```{r}
library(tibble)
library(tidyr)
library(ggplot2)
library(Rtsne)
library(umap)
library(Matrix)

set.seed(123)
source("./code/plot_ordering.R")
X <- as.matrix(sim$obs)
```

```{r}
library(smoothEMr)
set.seed(1234)
q <- 2
Q_prior_RW2 <- make_random_walk_precision(K = 65, 
                                          d=ncol(X), 
                                          lambda = 30, 
                                          q = q, 
                                          ridge = 0)
```


### Initialization using PCA

For comparison, let's also start with PCA initialization.

```{r}
init_params <- initialize_ordering(X = X, K = 65, method = "PCA")

result_RW2_PC1 <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 100,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

plot_EM_embedding(result_RW2_PC1, X, dims = c(1,2))
```


Just like what we would expect, the inferred curve is not able to recover the swiss-roll structure, as what we have seen in the principal curve fitting with PCA initialization.

### Initialization using Principal Curve

Now let's take the principal curve scores to initialize the SmoothEM.

```{r}
# pc_score <- pc_fit_lowess$lambda
# init_params <- make_init(X = X, ordering_vec = -pc_score, K = 65)
init_params <- initialize_ordering(X = X, K = 65, method = "pcurve")
result_RW2_pcurve <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 100,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)
plot_EM_embedding(result_RW2_pcurve, X, dims = c(1,2))
```


The principal curve initialization also fails to recover the swiss-roll structure.


### Initialization using t-SNE

Let's try t-SNE to get a 1D ordering of the data points for initialization.

```{r}
set.seed(123)
init_params <- initialize_ordering(X = X, K = 65, method = "tSNE")
result_RW2_tsne <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 100,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

plot_EM_embedding(result_RW2_tsne, X, dims = c(1,2))
```


The t-SNE initialization also fails to recover the swiss-roll structure.



### Initialization using Multi-scale SmoothEM

Now, let's try the progressive multi-scale SmoothEM initialization.

```{r fig.width=6, fig.height=4}
result_RW2_multi <- progressive_smoothEM(
  data = X,
  m_max = 6,              
  lambda_final = 100,
  q = 2,
  ridge = 0,
  nugget_kriging = 0,  
  tol = 1e-3,
  max_iter = 100,
  relative_lambda = T,
  plot_each_stage = F,
  verbose = F
)

plot_EM_embedding(result_RW2_multi$fits$K_65, X, dims = c(1,2))
```

Still no luck in recovering the swiss-roll structure.


## Initialization using Fiedler vector

Now, let's try a graph-based initialization using Fiedler vector.

```{r}
init_params <- initialize_ordering(
  X = X,
  K = 65, k = 20,
  method = "fiedler"
)
result_RW2_fiedler <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 100,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

plot_EM_embedding(result_RW2_fiedler, X, dims = c(1,2))
```

This time we are able to recover the swiss-roll structure quite well!


## Why many of these method do not work?

It seems like most of these methods do not provide a good visual recovery of the swiss-roll structure.

One possible reason is that the initialization orderings are all very different from the true underlying ordering (by t), and all of these iterative methods are really easy to get stuck in local optima in the ordering space.

Let's compare the ELBO traces from these different initialization to see how they behave.

```{r}
range_elbo <- c(
  min(c(result_RW2_PC1$elbo_trace,
        result_RW2_pcurve$elbo_trace,
        result_RW2_tsne$elbo_trace,
        result_RW2_multi$fits$K_65$elbo_trace,
        result_RW2_fiedler$elbo_trace)),
  max(c(result_RW2_PC1$elbo_trace,
        result_RW2_pcurve$elbo_trace,
        result_RW2_tsne$elbo_trace,
        result_RW2_multi$fits$K_65$elbo_trace,
        result_RW2_fiedler$elbo_trace))
)
# compare ELBOs
plot(NULL, xlim=c(1,100), ylim=range_elbo,
     xlab="Iteration", ylab="ELBO",
     main="ELBO Comparison")
lines(result_RW2_PC1$elbo_trace, col="blue", lwd=2)
lines(result_RW2_pcurve$elbo_trace, col="green", lwd=2)
lines(result_RW2_tsne$elbo_trace, col="purple", lwd=2)
lines(result_RW2_multi$fits$K_65$elbo_trace, col="orange", lwd=2)
lines(result_RW2_fiedler$elbo_trace, col="red", lwd=2)

legend("bottomright",
       legend=c("PCA", "Pcurve", "t-SNE", "Multi-scale", "Fiedler"),
       col=c("blue", "green", "purple", "orange", "red"),
       lwd=2)
```




