---
title: "Smooth EM algorithm with mixture of two latent orderings"
author: "Ziang Zhang"
date: "2025-07-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Problem Setup

We consider the following problem.

Let $\boldsymbol{x}_i \in \mathbb{R}^D$ denote observation $i$ (e.g., a cell), where $D$ is the total number of coordinates (e.g., genes).
The data matrix $\mathbf{X} \in \mathbb{R}^{N \times D}$ contains $N$ observations, each with $D$ coordinates.

We define a partition of the coordinates into two disjoint subsets:
$$
\mathcal{J}_1, \mathcal{J}_2 \subseteq \{1, \dots, D\}, \quad \mathcal{J}_1 \cup \mathcal{J}_2 = \{1, \dots, D\}, \quad \mathcal{J}_1 \cap \mathcal{J}_2 = \emptyset.
$$

Hence, each vector $\boldsymbol{x}_i$ can be decomposed as
$$
\boldsymbol{x}_i = \left( \boldsymbol{x}_i^{(1)}, \boldsymbol{x}_i^{(2)} \right),
$$
where 
$$
\boldsymbol{x}_i^{(1)} = (x_{ij})_{j \in \mathcal{J}_1} \in \mathbb{R}^{|\mathcal{J}_1|}, \quad 
\boldsymbol{x}_i^{(2)} = (x_{ij})_{j \in \mathcal{J}_2} \in \mathbb{R}^{|\mathcal{J}_2|}.
$$

For each partition $j \in \{1,2\}$, we assume a smooth latent structure such that
$$
\boldsymbol{x}_i^{(j)} \mid z_i^{(j)} \sim \mathcal{N}\left( \boldsymbol{\mu}^{(j)}(z_i^{(j)}),\; \boldsymbol{\Sigma}^{(j)} \right), \quad z_i^{(j)} \in \{1, \dots, K_j\}.
$$

Here, $z_i^{(j)}$ is a discretized latent index representing the position of observation $i$ along the latent ordering $j$. 
The function $\boldsymbol{\mu}^{(j)}(\cdot)$ defines a smooth mean curve over $K_j$ grid points. 
The covariance matrix $\boldsymbol{\Sigma}^{(j)}$ is typically assumed to follow a shared and diagonal structure (i.e., EEI) for simplicity.

Our target is to obtain the partitions $\mathcal{J}_1$ and $\mathcal{J}_2$, the latent orderings $z_i^{(1)}$ and $z_i^{(2)}$ (through each assignment matrix $\boldsymbol{\Gamma}^{(j)}$), and the smooth mean curves $\boldsymbol{\mu}^{(1)}(\cdot)$ and $\boldsymbol{\mu}^{(2)}(\cdot)$.


## An Iterative Algorithm

If the partition $\mathcal{J}_1$ and $\mathcal{J}_2$ are known, we can apply the standard smooth EM algorithm to estimate the parameters in each partition separately.  
However, the partition is unknown, and there are $2^D$ possible partitions, which is computationally infeasible to exhaustively search for large $D$.

Therefore, we propose the following iterative algorithm that jointly infers the partition and the model parameters in a greedy manner.

Let $\mathcal{J}_1^{(t)}$ and $\mathcal{J}_2^{(t)}$ denote the partition at iteration $t$. Each iteration then proceeds in three steps:

### Step 1: Coordinate Initialization

We initialize a partition of the coordinates into two groups, for example by random assignment:
\[
\mathcal{J}_1^{(0)} \cup \mathcal{J}_2^{(0)} = \{1, \dots, D\}, \quad \mathcal{J}_1^{(0)} \cap \mathcal{J}_2^{(0)} = \emptyset.
\]

### Step 2: Fit Two Smooth EM Models

Given the current partition $\mathcal{J}_1^{(t)}, \mathcal{J}_2^{(t)}$, we fit two separate smooth EM models to the data:

- For coordinates in $\mathcal{J}_1^{(t)}$, fit a $K_1$-component smooth mixture model and estimate $(\boldsymbol{\Gamma}^{(1)}, \boldsymbol{\mu}^{(1)}, \boldsymbol{\Sigma}^{(1)})$;
- For coordinates in $\mathcal{J}_2^{(t)}$, fit a $K_2$-component smooth mixture model and estimate $(\boldsymbol{\Gamma}^{(2)}, \boldsymbol{\mu}^{(2)}, \boldsymbol{\Sigma}^{(2)})$.

Each of these models can be solved using our current smooth EM algorithm, under a specified prior (e.g., random walk).

Here the most important quantities are $\boldsymbol{\Gamma}^{(j)}$, which essentially represent the latent positions of the observations along the latent ordering $j$.

### Step 3: Re-assign Coordinates via Posterior Optimization

Each coordinate $j \in \{1, \dots, D\}$ is re-assigned to either partition $\mathcal{J}_1$ or $\mathcal{J}_2$ based on the current position matrices $\boldsymbol{\Gamma}^{(1)}$ and $\boldsymbol{\Gamma}^{(2)}$, to optimize the log-posterior density of the data under each model.

Let $\boldsymbol{X}_{.j} \in \mathbb{R}^N$ denote the $j$th column of the data matrix.  
For each $j$, we compute the (negative) log-posterior density under each ordering:
\[
\mathcal{E}_j^{(1)} = \text{negative posterior under ordering 1}, \quad \mathcal{E}_j^{(2)} = \text{negative posterior under ordering 2}.
\]

Specifically, the negative log-posterior density for ordering $a$ is given by:
\[
\mathcal{E}_j^{(a)} = 
\underbrace{
\sum_{i=1}^N \sum_{k=1}^{K_a} \Gamma_{ik}^{(a)} 
\frac{\left(\mathbf{X}_{ij} - \mu_{jk}^{(a)}\right)^2}{2\, (\sigma_j^{(a)})^2}
}_{\text{data fit}}
\;+\;
\underbrace{
\sum_{i=1}^N \sum_{k=1}^{K_a} \Gamma_{ik}^{(a)} 
\log\left(\sigma_j^{(a)}\right)
}_{\text{obs. norm. const.}}
\;+\;
\underbrace{
\frac{1}{2} (\boldsymbol{\mu}_j^{(a)})^\top Q^{(a)} \boldsymbol{\mu}_j^{(a)}
}_{\text{smoothness penalty}}
\;+\;
\underbrace{
\frac{K_a}{2} \log(2\pi) - \frac{1}{2} \log \det(Q^{(a)})
}_{\text{prior norm. const.}}.
\]


We then update the assignment as follows:
\[
j \in \mathcal{J}_1^{(t+1)} \quad \text{if} \quad \mathcal{E}_j^{(1)} < \mathcal{E}_j^{(2)}, \quad 
j \in \mathcal{J}_2^{(t+1)} \quad \text{otherwise}.
\]

This procedure is iterated until convergence of the coordinate assignments.



## Simulation


```{r warning=FALSE, message=FALSE}
source("./code/prior_precision.R")
source("./code/general_EM.R")
source("./code/two_ordering_smoothEM.R")
library(tidyverse)
library(ggplot2)
library(reshape2)
library(tidyr)
library(dplyr)
library(MASS)
library(fields)  # for rdist and Matern
```


In this experiment, we generate synthetic data with **two distinct latent orderings**, each affecting a disjoint subset of the coordinates.

Specifically, we consider $D = 16$ coordinates, partitioned as follows:

- The first $8$ coordinates are generated as smooth functions of a latent variable $t^{(1)}$;
- The remaining $8$ coordinates are generated as smooth functions of another latent variable $t^{(2)}$.

The data matrix consists of $N = 1000$ observations. For each latent ordering $t^{(j)}$, we sample $t^{(j)}_i \sim \text{Uniform}(0, 1)$ independently for $i = 1, \dots, N$, and generate the smooth function in each coordinate as independent Matérn processes.

Below we construct the data:

```{r}
N <- 1000; D <- 16
set.seed(1234)
t1 <- runif(1000, min = 0, max = 1)
t2 <- runif(1000, min = 0, max = 1)

# GP kernel function using Matern
matern_cov <- function(t, range = 0.5, smoothness = 2.5, variance = 1.0) {
  dists <- rdist(t, t)
  cov_mat <- Matern(dists, range = range, smoothness = smoothness)
  cov_mat * variance
}

# Generate first 8 coordinates (dependent on t1)
Sigma1 <- matern_cov(matrix(t1, ncol = 1))
X1 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma1)
X1 <- t(X1)  # Make it N x 8

# Generate second 8 coordinates (dependent on t2)
Sigma2 <- matern_cov(matrix(t2, ncol = 1))
X2 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma2)
X2 <- t(X2)  # Make it N x 8

X <- cbind(X1, X2)  # Combine the two sets of coordinates
# permute the columns
permut_cols <- sample(1:16, 16)
X <- X[, permut_cols]  # Permute the columns of X
```


Let's visualize the generated data against the latent orderings $t^{(1)}$ and $t^{(2)}$.

```{r}
# Plot the first 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t1, X1[, i], 
       cex = 0.3,
       main = paste("Coordinate", i), 
       xlab = "t1", ylab = paste("X1[", i, "]"))
}
par(mfrow = c(1,1))

# Plot the second 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t2, X2[, i], 
       cex = 0.3,
       main = paste("Coordinate", i + 8), 
       xlab = "t2", ylab = paste("X2[", i, "]"))
}
par(mfrow = c(1,1))
```

Let's apply the smooth EM algorithm to infer the two latent orderings, with the second order random walk prior for $\boldsymbol{\mu}^{(1)}$ and $\boldsymbol{\mu}^{(2)}$.

```{r}
set.seed(123)
res <- two_ordering_smoothEM(
  X = X, K = 50,
  modelName = "EEI",
  make_Q_prior = make_random_walk_precision,
  lambda_vec = c(10, 10),
  relative_lambda = TRUE,
  rw_order_vec = c(2, 2),
  reassign_by = "posterior",
  max_iter = 100,
  verbose = TRUE
)
```

Let's check the inferred latent orderings.

```{r}
res$coord_assign[order(permut_cols)]
```

The first 8 coordinates are correctly assigned to one partition, and the second 8 coordinates are assigned to the other partition.


Let's visualize the inferred $\boldsymbol{\mu}^{(1)}$:

```{r}
par(mfrow = c(2, 4))
for (i in 1:length(res$result1$params$mu[[1]])) {
  mean_vec <- sapply(res$result1$params$mu, function(mu_k) mu_k[[i]])
  plot(1:50, mean_vec, 
       cex = 0.3,
       main = paste("Coordinate", i), 
       xlab = "discretized t1", ylab = paste("mu", i))
}
par(mfrow = c(1, 1))
```

Also visualize the inferred $\boldsymbol{\mu}^{(2)}$:

```{r}
par(mfrow = c(2, 4))
for (i in 1:length(res$result2$params$mu[[1]])) {
  mean_vec <- sapply(res$result2$params$mu, function(mu_k) mu_k[[i]])
  plot(1:50, mean_vec, 
       cex = 0.3,
       main = paste("Coordinate", i), 
       xlab = "discretized t2", ylab = paste("mu", i))
}
par(mfrow = c(1, 1))
```



