---
title: "explore_SmoothEM_Mixture_backward"
author: "Ziang Zhang"
date: "2025-12-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


---

## **Simulation**

```{r warning=FALSE, message=FALSE}
source("code/prior_precision.R")
source("code/general_EM_perCoordinate.R")
source("code/two_ordering_EM_backward.R")
source("code/plot_ordering.R")
library(tidyverse)
library(ggplot2)
library(reshape2)
library(tidyr)
library(dplyr)
library(MASS)
library(matrixStats) 
library(fields)  # for rdist and Matern
```

In this experiment, we generate synthetic data with **two distinct latent orderings**, each affecting a disjoint subset of the coordinates.

Specifically, we consider $D = 16$ coordinates, partitioned as follows:

- The first $8$ coordinates are generated as smooth functions of a latent variable $t^{(1)}$;
- The remaining $8$ coordinates are generated as smooth functions of another latent variable $t^{(2)}$.

The data matrix consists of $N = 1000$ observations. For each latent ordering $t^{(j)}$, we sample $t^{(j)}_i \sim \text{Uniform}(0, 10)$ independently for $i = 1, \dots, N$, and generate the smooth function in each coordinate as independent Matérn processes.

Below we construct the data:

```{r}
N <- 1000; D <- 16
set.seed(123)
t1 <- runif(1000, min = 0, max = 10)
t2 <- runif(1000, min = 0, max = 10)

# GP kernel function using Matern
matern_cov <- function(t, range = 5, smoothness = 2.5, variance = 3.0) {
  dists <- rdist(t, t)
  cov_mat <- Matern(dists, range = range, smoothness = smoothness)
  cov_mat * variance
}

# Generate first 8 coordinates (dependent on t1)
Sigma1 <- matern_cov(matrix(t1, ncol = 1))
X1 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma1)
X1 <- t(X1)  # Make it N x 8
# Make X1 positive
X1 <- X1 + abs(min(X1)) + 1  # Shift to make all values positive

# Generate second 8 coordinates (dependent on t2)
Sigma2 <- matern_cov(matrix(t2, ncol = 1))
X2 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma2)
X2 <- t(X2)  # Make it N x 8
# Make X2 positive
X2 <- X2 + abs(min(X2)) + 1  # Shift to make all values positive
```

(Optional:) Randomly make some coordinates thresholded to constant:
```{r eval=FALSE}
threshold_1 <- 4
threshold_2 <- 6
for (i in 1:ncol(X1)) {
  uu <- runif(1)
  if (uu < 0.4) {
    X1[order(t1),i][sort(t1) >= threshold_1] <- X1[order(t1),i][sort(t1) >= threshold_1][1]
  }
  else if(uu > 0.7){
    next
  }
  else{
    X1[order(t1),i][sort(t1) <= threshold_1] <- X1[order(t1),i][sort(t1) <= threshold_1][length(X1[order(t1),i][sort(t1) <= threshold_1])]
  }
}
for (i in 1:ncol(X2)) {
  uu <- runif(1)
  if (uu < 0.4) {
    X2[order(t2),i][sort(t2) >= threshold_2] <- X2[order(t2),i][sort(t2) >= threshold_2][1]
  }
  else if(uu > 0.7){
    next
  }
  else{
    X2[order(t2),i][sort(t2) <= threshold_2] <- X2[order(t2),i][sort(t2) <= threshold_2][length(X2[order(t2),i][sort(t2) <= threshold_2])]
  }
}
```

```{r}
X <- cbind(X1, X2)  # Combine the two sets of coordinates
# permute the columns
permut_cols <- sample(1:16, 16)
# permut_cols <- 1:16
X <- X[, permut_cols]  # Permute the columns of X

# Add some noise
X <- X + matrix(rnorm(N * D, mean = 0, sd = 0.05), nrow = N, ncol = D)
```

Let's visualize the generated data against the latent orderings $t^{(1)}$ and $t^{(2)}$.

```{r}
# Plot the first 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t1, X1[, i], 
       cex = 0.3,
       main = paste("Coordinate", i), 
       xlab = "t1", ylab = paste("X1[", i, "]"))
}
par(mfrow = c(1,1))

# Plot the second 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t2, X2[, i], 
       cex = 0.3,
       main = paste("Coordinate", i + 8), 
       xlab = "t2", ylab = paste("X2[", i, "]"))
}
par(mfrow = c(1,1))
```

Plot the coordinates in the same ordering in one figure:

```{r}
par(mfrow = c(1,2))
# Plot the first 8 coordinates together
plot(sort(t1), (X1[order(t1), 1] - mean(X1[ ,1])),
     cex = 0.3, ylim = c(-1,1),
     type = 'l',
     main = "First 8 Coordinates vs t1", 
     xlab = "t1", ylab = "X1")
for (i in 2:8) {
  lines(sort(t1), (X1[order(t1), i] - mean(X1[ ,i])), col = i)
}

# Plot the second 8 coordinates together
plot(sort(t2), (X2[order(t2), 1] - mean(X2[ ,1])),
     cex = 0.3, ylim = c(-1,1),
     type = 'l',
     main = "Second 8 Coordinates vs t2", 
     xlab = "t2", ylab = "X2")
for (i in 2:8) {
  lines(sort(t2), (X2[order(t2), i] - mean(X2[ ,i])), col = i)
}
par(mfrow = c(1,1))
```


Let's apply the smooth EM algorithm to infer the two latent orderings, with the second order random walk prior for $\boldsymbol{\mu}^{(1)}$ and $\boldsymbol{\mu}^{(2)}$.

```{r}
set.seed(1234)
K <- 50
Q1 <- make_random_walk_precision(K = K, d = 1, q = 2, lambda = 500)
Q2 <- make_random_walk_precision(K = K, d = 1, q = 2, lambda = 500)
res <- split_ordering_backward(
  X = X, K = K,
  init_method = "pca",
  Q1_K = Q1, Q2_K = Q2,
  rank_def1 = 2, rank_def2 = 2,
  verbose = 1
)
res$elbo1 + res$elbo2
```





## **Application to Pancreas dataset**

Now, let's see how this approach applies to the pancreas dataset, to distinguish between cell-type factors and batch-type factors.

```{r warning=FALSE, message=FALSE}
library(tibble)
library(Rtsne)
library(umap)
set.seed(1)
load("./data/loading_order/pancreas_factors.rdata")
load("./data/loading_order/pancreas.rdata")
cell_type_factors <- c(3, 8, 9, 12, 17, 18, 20, 21)
batch_type_factors <- c(1, 2, 4, 5, 6, 7, 11, 22)
all_factors <- c(cell_type_factors, batch_type_factors)
cells <- subsample_cell_types(sample_info$celltype,n = 500)
Loadings <- fl_snmf_ldf$L[cells,all_factors]
celltype <- as.character(sample_info$celltype[cells])
names(celltype) <- rownames(Loadings)
batchtype <- as.character(sample_info$tech[cells])
names(batchtype) <- rownames(Loadings)
```

Based on the visual assessment from [this analysis](https://stephenslab.github.io/single-cell-jamboree/pancreas_factors.html), we select the first 8 factors to be cell-type factors, and the remaining 8 factors to be batch-type factors.

Take a look at the unordered structure plot:

```{r}
plot_structure(Loadings)
```


Apply the smooth EM algorithm to infer the two latent orderings, with the second order random walk prior for $\boldsymbol{\mu}^{(1)}$ and $\boldsymbol{\mu}^{(2)}$.

Before we run the algorithm, we standardize the data by columns to ensure each loading factor has the same scale. 

```{r eval=FALSE}
# standardize by columns
Loadings <- scale(Loadings, center = FALSE, scale = apply(Loadings, 2, sd, na.rm = TRUE))
# Log1p
Loadings <- log1p(Loadings)
```

```{r}
set.seed(1234)
K <- 100
Q1 <- make_random_walk_precision(K = K, d = 1, q = 1, lambda = 500)
Q2 <- make_random_walk_precision(K = K, d = 1, q = 1, lambda = 500)
resPancreas <- split_ordering_backward(
  X = Loadings, K = K,
  init_method = "pca",
  Q1_K = Q1, Q2_K = Q2,
  rank_def1 = 1, rank_def2 = 1,
  verbose = 1
)
resPancreas$elbo1 + resPancreas$elbo2
```

Let's check the inferred latent orderings.

```{r}
resPancreas$J1
resPancreas$J2
```



Let's take a look at the structure plot ordered by each latent ordering.

```{r}
# position_1 <- apply(resPancreas$Gamma1, 1, which.max)
position_1 <- apply(resPancreas$Gamma1, 1, function(x) sum(x * seq(1, K)) / sum(x))

Loadings_J1 <- Loadings[, resPancreas$J1]
Loadings_J1 <- Loadings[, 5, drop = FALSE]
Loadings_J1_order <- order(position_1)
plot_structure(Loadings_J1, order = rownames(Loadings_J1)[Loadings_J1_order],
               plot_name = "Ordered by Latent Ordering 1")
plot_highlight_types(type_vec = celltype,
                     subset_types = unique(celltype),
                     order_vec = rownames(Loadings_J1)[Loadings_J1_order],
                     other_color = "white"
                     )
# plot_highlight_types(type_vec = batchtype,
#                      subset_types = unique(batchtype),
#                      order_vec = rownames(Loadings_J1)[Loadings_J1_order],
#                      other_color = "white"
#                      )
```


As well as the second latent ordering:

```{r}
# position_2 <- apply(resPancreas$Gamma2, 1, which.max)
position_2 <- apply(resPancreas$Gamma2, 1, function(x) sum(x * seq(1, K)) / sum(x))
Loadings_J2 <- Loadings[, resPancreas$J2, drop = FALSE]
Loadings_J2 <- Loadings[, 5, drop = FALSE]
Loadings_J2_order <- order(position_2)
plot_structure(Loadings_J2, order = rownames(Loadings_J2)[Loadings_J2_order],
               plot_name = "Ordered by Latent Ordering 2")
# plot_highlight_types(type_vec = batchtype,
#                      subset_types = unique(batchtype),
#                      order_vec = rownames(Loadings_J2)[Loadings_J2_order],
#                      other_color = "white"
#                      )
plot_highlight_types(type_vec = celltype,
                     subset_types = unique(celltype),
                     order_vec = rownames(Loadings_J2)[Loadings_J2_order],
                     other_color = "white"
                     )
```





