---
title: "Smooth EM in 2D (Additive)"
author: "Ziang Zhang"
date: "2025-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE}
source("./code/prior_precision.R")
source("./code/additive_EM.R")
source("./code/simulate2D_fields.R")
library(tidyverse)
library(ggplot2)
library(reshape2)
library(tidyr)
library(dplyr)
```

## Introduction

We consider a smooth EM algorithm where the component means $\mu(t_1,t_2)$ has an additive form as $\mu(t_1, t_2) = f(t_1) + g(t_2)$, where $f$ and $g$ are smooth functions of the 1D coordinates $t_1$ and $t_2$, respectively.

For simplicity, we assume $f$ and $g$ follow independent random walk priors, and are both defined on a equally spaced grid of $K$ points.

## Simulate Data

First, let's simulate some $d=16$-dimensional data from a $20 \times 20$ grid defined on the unit square $[0,1] \times [0,1]$. 

The observation noise is assumed to be Gaussian with standard deviation 0.1 in each dimension.

To make this problem more interesting, let's consider the sampling region follows roughly a donut shape.

```{r}
funcs <- make_diverse_additive_functions(n_surfaces = 16,
                                         freq_pool   = c(0.5, 1, 2),
                                         seed        = 123)

n           <- 100        # grid resolution
inner_rad   <- 0.2        # inner hole radius
outer_rad   <- 0.4        # outer radius of donut
center      <- c(0.5, 0.5)
xs  <- seq(0, 1, length.out = n)
ys  <- seq(0, 1, length.out = n)
grid <- expand.grid(x = xs, y = ys)
r <- sqrt((grid$x - center[1])^2 + (grid$y - center[2])^2)
grid_choice <- subset(grid, r > inner_rad & r < outer_rad)

# Quick plot to check
plot(grid, pch = 16, cex = 0.3, col = "lightgrey",
     main = "Donut shaped Grid")
points(grid_choice, pch = 16, cex = 0.5, col = "steelblue")

sim <- simulate_2d_processes(
  f          = funcs,
  nrow       = 50,
  ncol       = 50,
  d          = 16,
  epsilon_sd = 0.1,
  grid = grid_choice,
  seed       = 123
)
```

Let's visualize the simulated processes.

```{r}
df_plot <- cbind(sim$grid, sim$Y)
df_long <- melt(df_plot, id.vars = c("x", "y"),
                variable.name = "process",
                value.name = "value")

ggplot(df_long, aes(x = x, y = y, color = value)) +
  geom_point() +
  scale_color_viridis_c(name = "value") +
  coord_equal() +
  facet_wrap(~ process, nrow = 4) +
  labs(
    title = "Simulated 2D Processes (colored dots)",
    x = "x",
    y = "y"
  ) +
  theme_minimal()
```

Now, let's permute the rows of `sim$Y` to avoid any spatial structure in the ordering of the data points.

```{r}
sim$Y <- sim$Y[sample(nrow(sim$Y)), ]
```


## Fitting RW-2

We will fit the additive smooth EM algorithm using RW2 priors for both $f$ and $g$.

```{r}
K <- 20
Q1 <- make_random_walk_precision(K = K, d = ncol(sim$Y), q = 2, lambda = 1e4, ridge = 0)
Q2 <- make_random_walk_precision(K = K, d = ncol(sim$Y), q = 2, lambda = 1e4, ridge = 1e-10)
Q_list <- list(Q1, Q2)
set.seed(123)
init_params <- make_default_init_add(X = sim$Y, Q_list = Q_list)
result_RW <- additive_smoothEM(
  data = sim$Y,
  Q_list = Q_list,
  init_params = init_params,
  max_iter = 100,
  tol = 1e-4,
  relative_lambda = TRUE,
  verbose = TRUE
)
plot(result_RW$loglik_trace)
```


```{r}
mu_list <- result_RW$params$mu
n_cells <- length(mu_list)   
K       <- sqrt(n_cells)     

df <- do.call(rbind, lapply(seq_along(mu_list), function(i) {
  row <- ((i - 1) %/% K) + 1
  col <- ((i - 1) %%  K) + 1
  vals <- mu_list[[i]]
  data.frame(
    row = row,
    col = col,
    dim = factor(1:length(vals)), 
    value = vals
  )
}))

df$col <- df$col/K
df$row <- df$row/K

ggplot(df, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D") +
  facet_wrap(~ dim, nrow = 4,
             labeller = labeller(dim = function(x) paste("Dimension", x))) +
  coord_equal() +
  labs(
    x = "column index",
    y = "row index",
    fill = "mu value",
    title = "RW: Estimated mu values on lattice"
  ) +
  theme_minimal()
```

Plot each function:

```{r}
par(mfrow = c(4, 4), mar = c(2, 2, 2, 1))
for (i in 1:ncol(result_RW$theta[[1]])) {
  plot(result_RW$theta[[1]][,i], type = 'l', main = paste("Coordinate", i))
}
par(mfrow = c(1, 1))

par(mfrow = c(4, 4), mar = c(2, 2, 2, 1))
for (i in 1:ncol(result_RW$theta[[2]])) {
  plot(result_RW$theta[[2]][,i], type = 'l', main = paste("Coordinate", i))
}
par(mfrow = c(1, 1))
```

Let's take a look at the inferred locations:

```{r}
K1 <- K; K2 <- K

# 1) Hard‐assign each observation to a latent grid index (1 … K1*K2)
cluster_assign <- max.col(result_RW$gamma)   

# 2) Build a data.frame of the full latent grid with its (i,j) coordinates
latent_grid <- expand.grid(i = seq_len(K1), j = seq_len(K2))
latent_grid$cluster <- seq_len(nrow(latent_grid))  # 1:(K1*K2)

# 3) Map each observation to its latent grid (i,j)
obs_map <- data.frame(
  obs     = seq_len(nrow(sim$Y)),
  cluster = cluster_assign
)
assign_df <- merge(obs_map, latent_grid, by = "cluster")

# 4) Compute raw counts per (i,j)
raw_cnt <- assign_df %>%
  count(i, j, name = "n_obs")

# 5) Build the complete grid of (i,j)
full_grid <- expand.grid(i = seq_len(K1), j = seq_len(K2))

# 6) Left‐join counts and fill missing with zero
cnt_full <- full_grid %>%
  left_join(raw_cnt, by = c("i", "j")) %>%
  replace_na(list(n_obs = 0))

# 7) Plot the number of observations per latent grid cell,
#    normalized to [0,1] for both axes
ggplot(cnt_full, aes(x = j / K2, y = i / K1, fill = n_obs)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D", name = "# obs") +
  labs(
    x     = "latent j / K2",
    y     = "latent i / K1",
    title = "Observations per Latent Grid Cell"
  ) +
  coord_equal() +
  theme_minimal()
```


Thought the inferred locations are not perfectly matching the donut shape, they still show some "donut-like" structure.
