---
title: "Smooth EM algorithm with mixture of two latent orderings"
author: "Ziang Zhang"
date: "2025-07-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## **Problem Setup**

We consider the following problem.

Let $\boldsymbol{x}_i \in \mathbb{R}^D$ denote observation $i$ (e.g., a cell), where $D$ is the total number of coordinates (e.g., genes).
The data matrix $\mathbf{X} \in \mathbb{R}^{N \times D}$ contains $N$ observations, each with $D$ coordinates.

We define a partition of the coordinates into two disjoint subsets:
$$
\mathcal{J}_1, \mathcal{J}_2 \subseteq \{1, \dots, D\}, \quad \mathcal{J}_1 \cup \mathcal{J}_2 = \{1, \dots, D\}, \quad \mathcal{J}_1 \cap \mathcal{J}_2 = \emptyset.
$$

Hence, each vector $\boldsymbol{x}_i$ can be decomposed as
$$
\boldsymbol{x}_i = \left( \boldsymbol{x}_i^{(1)}, \boldsymbol{x}_i^{(2)} \right),
$$
where 
$$
\boldsymbol{x}_i^{(1)} = (x_{ij})_{j \in \mathcal{J}_1} \in \mathbb{R}^{|\mathcal{J}_1|}, \quad 
\boldsymbol{x}_i^{(2)} = (x_{ij})_{j \in \mathcal{J}_2} \in \mathbb{R}^{|\mathcal{J}_2|}.
$$

For each partition $j \in \{1,2\}$, we assume a smooth latent structure such that
$$
\boldsymbol{x}_i^{(j)} \mid z_i^{(j)} \sim \mathcal{N}\left( \boldsymbol{\mu}^{(j)}(z_i^{(j)}),\; \boldsymbol{\Sigma}^{(j)} \right), \quad z_i^{(j)} \in \{1, \dots, K_j\}.
$$

Here, $z_i^{(j)}$ is a discretized latent index representing the position of observation $i$ along the latent ordering $j$. 
The function $\boldsymbol{\mu}^{(j)}(\cdot)$ defines a smooth mean curve over $K_j$ grid points. 
The covariance matrix $\boldsymbol{\Sigma}^{(j)}$ is typically assumed to follow a shared and diagonal structure (i.e., EEI) for simplicity.

Our target is to obtain the partitions $\mathcal{J}_1$ and $\mathcal{J}_2$, the latent orderings $z_i^{(1)}$ and $z_i^{(2)}$ (through each assignment matrix $\boldsymbol{\Gamma}^{(j)}$), and the smooth mean curves $\boldsymbol{\mu}^{(1)}(\cdot)$ and $\boldsymbol{\mu}^{(2)}(\cdot)$.


## **An Iterative Algorithm**

If the partition $\mathcal{J}_1$ and $\mathcal{J}_2$ are known, we can apply the standard smooth EM algorithm to estimate the parameters in each partition separately. However, the partition is unknown, and there are $2^D$ possible partitions, which is computationally infeasible to exhaustively search for large $D$. Therefore, we propose the following iterative algorithm that jointly infers the partition and the model parameters in a greedy manner.

Let $\mathcal{J}_1^{(t)}$ and $\mathcal{J}_2^{(t)}$ denote the partition at iteration $t$. Each iteration then proceeds in three steps:

### Step 1: Coordinate Initialization

We initialize a partition of the coordinates into two groups, for example by random assignment:
\[
\mathcal{J}_1^{(0)} \cup \mathcal{J}_2^{(0)} = \{1, \dots, D\}, \quad \mathcal{J}_1^{(0)} \cap \mathcal{J}_2^{(0)} = \emptyset.
\]

### Step 2: Fit Two Smooth EM Models

Given the current partition $\mathcal{J}_1^{(t)}, \mathcal{J}_2^{(t)}$, we fit two separate smooth EM models to the data:

- For coordinates in $\mathcal{J}_1^{(t)}$, fit a $K_1$-component smooth mixture model and estimate $(\boldsymbol{\Gamma}^{(1)}, \boldsymbol{\mu}^{(1)}, \boldsymbol{\Sigma}^{(1)})$;
- For coordinates in $\mathcal{J}_2^{(t)}$, fit a $K_2$-component smooth mixture model and estimate $(\boldsymbol{\Gamma}^{(2)}, \boldsymbol{\mu}^{(2)}, \boldsymbol{\Sigma}^{(2)})$.

Each of these models can be solved using our current smooth EM algorithm, under a specified prior (e.g., random walk).

Here the most important quantities are $\boldsymbol{\Gamma}^{(j)}$, which essentially represent the latent positions of the observations along the latent ordering $j$.

### Step 3: Re-assign Coordinates via Posterior Optimization

Each coordinate $j \in \{1, \dots, D\}$ is re-assigned to either partition $\mathcal{J}_1$ or $\mathcal{J}_2$ based on the current position matrices $\boldsymbol{\Gamma}^{(1)}$ and $\boldsymbol{\Gamma}^{(2)}$, to optimize the log-posterior density of the data under each model.

Let $\boldsymbol{X}_{.j} \in \mathbb{R}^N$ denote the $j$th column of the data matrix. For each $j$, we compute the (negative) log-posterior density under each ordering:
\[
\mathcal{E}_j^{(1)} = \text{negative posterior under ordering 1}, \quad \mathcal{E}_j^{(2)} = \text{negative posterior under ordering 2}.
\]

Specifically, the negative log-posterior density for ordering $a$ is given by:
\[
\mathcal{E}_j^{(a)} = 
\underbrace{
\sum_{i=1}^N \sum_{k=1}^{K_a} \Gamma_{ik}^{(a)} 
\frac{\left(\mathbf{X}_{ij} - \mu_{jk}^{(a)}\right)^2}{2\, (\sigma_j^{(a)})^2}
}_{\text{data fit}}
\;+\;
\underbrace{
\sum_{i=1}^N \sum_{k=1}^{K_a} \Gamma_{ik}^{(a)} 
\log\left(\sigma_j^{(a)}\right)
}_{\text{obs. norm. const.}}
\;+\;
\underbrace{
\frac{1}{2} (\boldsymbol{\mu}_j^{(a)})^\top Q^{(a)} \boldsymbol{\mu}_j^{(a)}
}_{\text{smoothness penalty}}
\;+\;
\underbrace{
\frac{K_a}{2} \log(2\pi) - \frac{1}{2} \log \det(Q^{(a)})
}_{\text{prior norm. const.}}.
\]


We then update the assignment as follows:
\[
j \in \mathcal{J}_1^{(t+1)} \quad \text{if} \quad \mathcal{E}_j^{(1)} < \mathcal{E}_j^{(2)}, \quad 
j \in \mathcal{J}_2^{(t+1)} \quad \text{otherwise}.
\]

This procedure is iterated until convergence of the coordinate assignments.



## **Simulation**


```{r warning=FALSE, message=FALSE}
source("./code/prior_precision.R")
source("./code/general_EM.R")
source("./code/two_ordering_smoothEM.R")
source("./code/plot_ordering.R")
library(tidyverse)
library(ggplot2)
library(reshape2)
library(tidyr)
library(dplyr)
library(MASS)
library(fields)  # for rdist and Matern
```


In this experiment, we generate synthetic data with **two distinct latent orderings**, each affecting a disjoint subset of the coordinates.

Specifically, we consider $D = 16$ coordinates, partitioned as follows:

- The first $8$ coordinates are generated as smooth functions of a latent variable $t^{(1)}$;
- The remaining $8$ coordinates are generated as smooth functions of another latent variable $t^{(2)}$.

The data matrix consists of $N = 1000$ observations. For each latent ordering $t^{(j)}$, we sample $t^{(j)}_i \sim \text{Uniform}(0, 10)$ independently for $i = 1, \dots, N$, and generate the smooth function in each coordinate as independent Matérn processes.

Below we construct the data:

```{r}
N <- 1000; D <- 16
set.seed(1234)
t1 <- runif(1000, min = 0, max = 10)
t2 <- runif(1000, min = 0, max = 10)

# GP kernel function using Matern
matern_cov <- function(t, range = 5, smoothness = 2.5, variance = 3.0) {
  dists <- rdist(t, t)
  cov_mat <- Matern(dists, range = range, smoothness = smoothness)
  cov_mat * variance
}

# Generate first 8 coordinates (dependent on t1)
Sigma1 <- matern_cov(matrix(t1, ncol = 1))
X1 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma1)
X1 <- t(X1)  # Make it N x 8
# Make X1 positive
X1 <- X1 + abs(min(X1)) + 1  # Shift to make all values positive

# Generate second 8 coordinates (dependent on t2)
Sigma2 <- matern_cov(matrix(t2, ncol = 1))
X2 <- mvrnorm(n = 8, mu = rep(0, N), Sigma = Sigma2)
X2 <- t(X2)  # Make it N x 8
# Make X2 positive
X2 <- X2 + abs(min(X2)) + 1  # Shift to make all values positive

X <- cbind(X1, X2)  # Combine the two sets of coordinates
# permute the columns
permut_cols <- sample(1:16, 16)
X <- X[, permut_cols]  # Permute the columns of X

# Add some noise
X <- X + matrix(rnorm(N * D, mean = 0, sd = 0.05), nrow = N, ncol = D)
```


Let's visualize the generated data against the latent orderings $t^{(1)}$ and $t^{(2)}$.

```{r}
# Plot the first 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t1, X1[, i], 
       cex = 0.3,
       main = paste("Coordinate", i), 
       xlab = "t1", ylab = paste("X1[", i, "]"))
}
par(mfrow = c(1,1))

# Plot the second 8 coordinates
par(mfrow = c(2,4))
for (i in 1:8) {
  plot(t2, X2[, i], 
       cex = 0.3,
       main = paste("Coordinate", i + 8), 
       xlab = "t2", ylab = paste("X2[", i, "]"))
}
par(mfrow = c(1,1))
```

Let's apply the smooth EM algorithm to infer the two latent orderings, with the second order random walk prior for $\boldsymbol{\mu}^{(1)}$ and $\boldsymbol{\mu}^{(2)}$.

```{r}
set.seed(123)
res <- two_ordering_smoothEM(
  X = X, K = 50,
  modelName = "EEI",
  init_method = "pca",
  init_coord_assign = rep(c(1,2), each = ncol(X) / 2),
  make_Q_prior = make_random_walk_precision,
  lambda_vec = c(50, 50),
  relative_lambda = TRUE,
  rw_order_vec = c(2, 2),
  reassign_by = "posterior",
  max_iter = 100,
  max_iter_EM = 200,
  verbose = TRUE
)
```

Let's check the inferred latent orderings.

```{r}
sort(permut_cols[res$J1])
sort(permut_cols[res$J2])
```

The first 8 coordinates are correctly assigned to one partition, and the second 8 coordinates are assigned to the other partition.


Let's visualize the inferred $\boldsymbol{\mu}^{(1)}$:

```{r}
all_coordinates_1 <- permut_cols[res$J1]
par(mfrow = c(2, 4))
for (i in 1:length(res$result1$params$mu[[1]])) {
  index <- order(all_coordinates_1)[i]
  mean_vec <- sapply(res$result1$params$mu, function(mu_k) mu_k[[index]])
  plot((1:50)/5, mean_vec, type = 'o',
       cex = 0.5,
       main = paste("Coordinate", all_coordinates_1[index]), 
       xlab = "discretized t1", ylab = paste("mu"))
  points(t1, 
         X[, permut_cols == all_coordinates_1[index]], 
         col = rgb(1, 0.75, 0.8, alpha = 0.8),
         cex = 0.2)
}
par(mfrow = c(1, 1))
```

Also visualize the inferred $\boldsymbol{\mu}^{(2)}$:

```{r}
all_coordinates_2 <- permut_cols[res$J2]
par(mfrow = c(2, 4))
for (i in 1:length(res$result2$params$mu[[1]])) {
  index <- order(all_coordinates_2)[i]
  mean_vec <- sapply(res$result2$params$mu, function(mu_k) mu_k[[index]])
  plot((1:50)/5, mean_vec, type = 'o',
       cex = 0.5,
       main = paste("Coordinate", all_coordinates_2[index]), 
       xlab = "discretized t2", ylab = paste("mu"))
  
  points((10 - t2), type = 'p',
         X[, permut_cols == all_coordinates_2[index]], 
         col = rgb(1, 0.75, 0.8, alpha = 0.8),
         cex = 0.2)
}
par(mfrow = c(1, 1))
```

Finally, let's try to examine the structure plot. To start with, we will take a look at the unordered data matrix.

```{r}
rownames(X) <- paste0("Cell", 1:nrow(X))
plot_structure(X)
```

Now, let's order them based on the first latent ordering:

```{r}
position_1 <- apply(res$Gamma1, 1, which.max)
X_J1 <- X[, res$J1]
X_J1_order <- order(position_1)
plot_structure(X_J1, order = rownames(X_J1)[X_J1_order])
```

As well as the second latent ordering:

```{r}
position_2 <- apply(res$Gamma2, 1, which.max)
X_J2 <- X[, res$J2]
X_J2_order <- order(position_2)
plot_structure(X_J2, order = rownames(X_J2)[X_J2_order])
```



## **Application to Pancreas dataset**

Now, let's see how this approach applies to the pancreas dataset, to distinguish between cell-type factors and batch-type factors.

```{r warning=FALSE, message=FALSE}
library(tibble)
library(Rtsne)
library(umap)
set.seed(1)
load("./data/loading_order/pancreas_factors.rdata")
load("./data/loading_order/pancreas.rdata")
cell_type_factors <- c(3,8,9,12,17,18,20,21)
batch_type_factors <- c(1, 2, 4, 5, 6, 7, 11, 22)
all_factors <- c(cell_type_factors, batch_type_factors)
cells <- subsample_cell_types(sample_info$celltype,n = 500)
Loadings <- fl_snmf_ldf$L[cells,all_factors]
celltype <- as.character(sample_info$celltype[cells])
names(celltype) <- rownames(Loadings)
batchtype <- as.character(sample_info$tech[cells])
names(batchtype) <- rownames(Loadings)
```

Based on the visual assessment from [this analysis](https://stephenslab.github.io/single-cell-jamboree/pancreas_factors.html), we select the first 8 factors to be cell-type factors, and the remaining 8 factors to be batch-type factors.

Let's permute the columns of the loading matrix to make it unordered:

```{r}
permut_cols <- sample(1:ncol(Loadings), ncol(Loadings))
Loadings <- Loadings[, permut_cols]
```

Take a look at the unordered structure plot:

```{r}
plot_structure(Loadings)
```


Apply the smooth EM algorithm to infer the two latent orderings, with the second order random walk prior for $\boldsymbol{\mu}^{(1)}$ and $\boldsymbol{\mu}^{(2)}$.

```{r}
resPancreas <- two_ordering_smoothEM(
  X = Loadings, K = 50,
  init_method = "pca",
  init_coord_assign = rep(c(1,2), each = ncol(Loadings) / 2),
  modelName = "EEI",
  make_Q_prior = make_random_walk_precision,
  lambda_vec = c(50, 50),
  relative_lambda = TRUE,
  rw_order_vec = c(2, 2),
  reassign_by = "posterior",
  max_iter = 100,
  max_iter_EM = 1,
  verbose = TRUE
)
```

Let's check the inferred latent orderings.

```{r}
sort(permut_cols[resPancreas$J1])
sort(permut_cols[resPancreas$J2])
```


Let's take a look at the structure plot ordered by each latent ordering.

```{r}
position_1 <- apply(resPancreas$Gamma1, 1, which.max)
Loadings_J1 <- Loadings[, resPancreas$J1]
Loadings_J1_order <- order(position_1)
plot_structure(Loadings_J1, order = rownames(Loadings_J1)[Loadings_J1_order],
               plot_name = "Ordered by Latent Ordering 1")
plot_highlight_types(type_vec = batchtype,
                     subset_types = unique(batchtype),
                     order_vec = rownames(Loadings_J1)[Loadings_J1_order],
                     other_color = "white"
                     )
```


As well as the second latent ordering:

```{r}
position_2 <- apply(resPancreas$Gamma2, 1, which.max)
Loadings_J2 <- Loadings[, resPancreas$J2]
Loadings_J2_order <- order(position_2)
plot_structure(Loadings_J2, order = rownames(Loadings_J2)[Loadings_J2_order],
               plot_name = "Ordered by Latent Ordering 2")
plot_highlight_types(type_vec = celltype,
                     subset_types = unique(celltype),
                     order_vec = rownames(Loadings_J2)[Loadings_J2_order],
                     other_color = "white"
                     )
```



