---
title: "Smooth EM in 2D"
author: "Ziang Zhang"
date: "2025-07-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We consider a smooth EM algorithm where the component means $\boldsymbol{\mu}_{1,1}, \dots, \boldsymbol{\mu}_{K,K}$ are placed on a 2D equally spaced lattice of size $K \times K$. 
To encourage **spatial smoothness**, we impose a **random walk of order $q$ (RW-$q$)** prior on the means.

Let $\boldsymbol{\mu}$ denote the stacked vector of all component means. The prior takes the form:

$$
p(\boldsymbol{\mu}) \propto \exp\left(-\frac{\lambda}{2} \boldsymbol{\mu}^\top Q \boldsymbol{\mu}\right),
$$

where $Q$ is a sparse precision matrix encoding $q$-th order discrete differences across the lattice.

For example, under a first-order random walk (RW-1), the penalty encourages each $\mu_{i,j}$ to be close to the average of its immediate neighbors:

$$
\mu_{i,j} \approx \frac{1}{4} (\mu_{i+1,j} + \mu_{i-1,j} + \mu_{i,j+1} + \mu_{i,j-1}).
$$

This model is sometimes called a Intrinsic Conditional AutoRegressive (ICAR) model in spatial statistics.

```{r}
library(ggplot2)
library(reshape2)
source("./code/prior_precision.R")
source("./code/general_EM.R")
source("./code/simulate2D_fields.R")
```

## Simulate Data

First, let's simulate some $d=16$-dimensional data from a $20 \times 20$ grid defined on the unit square $[0,1] \times [0,1]$. 
Each dimension is a smooth function of the 2D coordinates $(x,y)$.
The smooth function is randomly simulated from a Matern process with smoothness parameter $\nu=2$ and range parameter $\rho=0.3$.

The observation noise is assumed to be Gaussian with standard deviation 0.1 in each dimension.

```{r}
sim <- simulate_2d_processes(
  f          = NULL,
  nrow       = 20,
  ncol       = 20,
  d          = 16,
  epsilon_sd = 0.1,
  seed       = 123
)
```

Let's visualize the simulated processes.

```{r}
df_plot <- cbind(sim$grid, sim$Y)
df_long <- melt(df_plot, id.vars = c("x", "y"),
                variable.name = "process",
                value.name = "value")

ggplot(df_long, aes(x = x, y = y, color = value)) +
  geom_point() +
  scale_color_viridis_c(name = "value") +
  coord_equal() +
  facet_wrap(~ process, nrow = 4) +
  labs(
    title = "Simulated 2D Processes (colored dots)",
    x = "x",
    y = "y"
  ) +
  theme_minimal()
```

Now, let's permute the rows of `sim$Y` to avoid any spatial structure in the ordering of the data points.

```{r}
sim$Y <- sim$Y[sample(nrow(sim$Y)), ]
```


## Fitting RW-1

First, we fit the model with a first-order random walk (RW-1) prior, with $K=10$ (i.e., a $10 \times 10$ lattice of components). 

```{r}
K <- 10
Q_prior_RW1 <- make_lattice_rwq_precision(K=K, d = ncol(sim$Y), q = 1, lambda = 2, ridge = 0)
set.seed(123)
init_params <- make_default_init(sim$Y, K=(K^2))

result_RW1 <- EM_algorithm(
  data = sim$Y,
  Q_prior = Q_prior_RW1,
  init_params = init_params,
  max_iter = 200,
  modelName = "EII",
  relative_lambda = T,
  tol = 1e-2,
  verbose = TRUE
)
```

Let's visualize the smoothed component means $\boldsymbol{\mu}_{i,j}$ inferred for each dimension.

```{r}
mu_list <- result_RW1$params$mu
n_cells <- length(mu_list)   
K       <- sqrt(n_cells)     
df <- do.call(rbind, lapply(seq_along(mu_list), function(i) {
  row <- ((i - 1) %/% K) + 1
  col <- ((i - 1) %%  K) + 1
  vals <- mu_list[[i]]
  data.frame(
    row = row,
    col = col,
    dim = factor(1:length(vals)), 
    value = vals
  )
}))

df$col <- df$col/K
df$row <- df$row/K

ggplot(df, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D") +
  facet_wrap(~ dim, nrow = 4,
             labeller = labeller(dim = function(x) paste("Dimension", x))) +
  coord_equal() +
  labs(
    x = "column index",
    y = "row index",
    fill = "mu value",
    title = "RW1: Estimated mu values on 5×5 lattice"
  ) +
  theme_minimal()
```




## Fitting RW-2

Next, we fit the model with a second-order random walk (RW-2) prior, again with $K=10$.

```{r}
K <- 10
Q_prior_RW2 <- make_lattice_rwq_precision(K=K, d = ncol(sim$Y), q = 2, lambda = 1, ridge = 0)
set.seed(123)
init_params <- make_default_init(sim$Y, K=(K^2))

result_RW2 <- EM_algorithm(
  data = sim$Y,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 200,
  modelName = "EII",
  relative_lambda = TRUE,
  tol = 1e-2,
  verbose = TRUE
)
```

Let's visualize the smoothed component means $\boldsymbol{\mu}_{i,j}$ inferred for each dimension.

```{r}
mu_list <- result_RW2$params$mu
n_cells <- length(mu_list)   
K       <- sqrt(n_cells)     

df <- do.call(rbind, lapply(seq_along(mu_list), function(i) {
  row <- ((i - 1) %/% K) + 1
  col <- ((i - 1) %%  K) + 1
  vals <- mu_list[[i]]
  data.frame(
    row = row,
    col = col,
    dim = factor(1:length(vals)), 
    value = vals
  )
}))

df$col <- df$col/K
df$row <- df$row/K

ggplot(df, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D") +
  facet_wrap(~ dim, nrow = 4,
             labeller = labeller(dim = function(x) paste("Dimension", x))) +
  coord_equal() +
  labs(
    x = "column index",
    y = "row index",
    fill = "mu value",
    title = "RW2: Estimated mu values on 5×5 lattice"
  ) +
  theme_minimal()
```



## Fitting PCA

As a comparison, let's fit PCA to the data and visualize how each coordinate varies over the PC1–PC2 plane.

```{r}
pca_res <- prcomp(sim$Y, center = TRUE, scale. = FALSE)
# 1. Prepare a data frame with PC1, PC2, and each coordinate’s values
d  <- ncol(sim$Y)
df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  sim$Y
)
colnames(df)[3:(2+d)] <- paste0("coord", seq_len(d))

# 2. Melt to long format
df_long <- melt(df,
                id.vars       = c("PC1","PC2"),
                variable.name = "coordinate",
                value.name    = "value")

# 3. Plot a “surface” of mean coordinate value over the PC1–PC2 plane
#    We bin the PC1/PC2 space and compute the average 'value' in each bin.
ggplot(df_long, aes(x = PC1, y = PC2, z = value)) +
  stat_summary_2d(fun = mean, bins = 30) +
  scale_fill_viridis_c(name = "mean value") +
  facet_wrap(~ coordinate, nrow = 4) +
  coord_equal() +
  labs(
    title = "Coordinate surfaces over PC1–PC2",
    x     = "PC1",
    y     = "PC2"
  ) +
  theme_minimal()
```









