---
title: "Explore different ways of initializing SmoothEM"
author: "Ziang Zhang"
date: "2025-12-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We will consider implementing the SmoothEM with the RW2 prior for the pancrea dataset, and compare different ways of initializing the SmoothEM algorithm: 

```{r message = FALSE}
library(tibble)
library(tidyr)
library(ggplot2)
library(Rtsne)
library(umap)
library(princurve)

set.seed(123)
source("./code/plot_ordering.R")
source("./code/linear_EM.R")
source("./code/general_EM.R")
source("./code/prior_precision.R")
source("./code/kriging.R")
load("./data/loading_order/pancreas_factors.rdata")
load("./data/loading_order/pancreas.rdata")
cells <- subsample_cell_types(sample_info$celltype,n = 500)
Loadings <- fl_snmf_ldf$L[cells,c(3,8,9,12,17,18,20,21)]
# Loadings <- fl_snmf_ldf$L[cells,]
celltype <- as.character(sample_info$celltype[cells])
names(celltype) <- rownames(Loadings)
batchtype <- as.character(sample_info$tech[cells])
names(batchtype) <- rownames(Loadings)

# Let's further restrict cells to only contain cells from one type of batch
Loadings <- Loadings[batchtype == "inDrop3",]
celltype <- celltype[batchtype == "inDrop3"]
batchtype <- batchtype[batchtype == "inDrop3"]
```

```{r}
set.seed(1234)
q <- 2
Q_prior_RW2 <- make_random_walk_precision(K = 65, 
                                          d=ncol(Loadings), 
                                          lambda = 300, 
                                          q = q, 
                                          ridge = 0)
```


## Initialization with PC

First, we consider the SmoothEM initialized with the first principal component.

The original structure plot ordered by first PC:

```{r fig.width=6, fig.height=4}
PC1 <- prcomp(Loadings,center = TRUE, scale. = FALSE)$x[,1]
loadings_order_PC1 <- order(-PC1)
plot_structure(Loadings, order = rownames(Loadings)[loadings_order_PC1])
```

The smoothEM initialized with first PC:

```{r fig.width=6, fig.height=4}
init_params <- make_init(X = Loadings, ordering_vec = -PC1, K = 65)
result_RW2_PC1 <- EM_algorithm(
  data = Loadings,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 50,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(Loadings),
  nugget = 0,
  relative_lambda = TRUE,
  verbose = F
)
result_RW2_PC1$clustering <- apply(result_RW2_PC1$gamma, 1, which.max)
result_RW2_PC1$position <- result_RW2_PC1$gamma %*% matrix(1:65, ncol=1)
loadings_order_RW2_PC1 <- order(result_RW2_PC1$position)
plot_structure(Loadings, order = rownames(Loadings)[loadings_order_RW2_PC1])
```


## Initialization with P-curve

Now, we will consider initializing the SmoothEM with the principal curve.

The structure plot ordered by principal curve:
```{r fig.width=6, fig.height=4}
pc_fit <- principal_curve(as.matrix(Loadings), maxit = 1000, smoother = "smooth_spline", thresh = 1e-5)
pc_score <- pc_fit$lambda
loadings_order_pcurve <- order(-pc_score)
plot_structure(Loadings, order = rownames(Loadings)[loadings_order_pcurve])
```

The smoothEM initialized with principal curve:

```{r fig.width=6, fig.height=4}
init_params <- make_init(X = Loadings, ordering_vec = -pc_score, K = 65)
result_RW2_pcurve <- EM_algorithm(
  data = Loadings,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 50,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(Loadings),
  nugget = 0,
  relative_lambda = TRUE,
  verbose = F
)
result_RW2_pcurve$clustering <- apply(result_RW2_pcurve$gamma, 1, which.max)
result_RW2_pcurve$position <- result_RW2_pcurve$gamma %*% matrix(1:65, ncol=1)
loadings_order_RW2_pcurve <- order(result_RW2_pcurve$position)
plot_structure(Loadings, order = rownames(Loadings)[loadings_order_RW2_pcurve])
```

## Initialization with multi-scale

Finally, we will consider initializing the SmoothEM with the multi-scale progressive smoothing approach.

We start with a small choice of $K$ (e.g., K=2), and gradually increase $K$ to the desired value (e.g., K=65) by equally splitting the intervals between two grid-locations.
The mean values from the previous stage will be kriged to get the initial mean values for the next stage.

```{r fig.width=6, fig.height=4}
result_RW2_multi <- progressive_smoothEM(
  data = Loadings,
  m_max = 6,              
  lambda_final = 300,
  q = 2,
  ridge = 0,
  nugget_kriging = 0,  
  tol = 1e-3,
  max_iter = 50,
  relative_lambda = TRUE,
  plot_each_stage = F,
  verbose = F
)

result_RW2_multi$clustering <- apply(result_RW2_multi$fits$K_65$gamma, 1, which.max)
result_RW2_multi$position <- result_RW2_multi$fits$K_65$gamma %*% matrix(1:65, ncol=1)
loadings_order_RW2_multi <- order(result_RW2_multi$position)
plot_structure(Loadings, order = rownames(Loadings)[loadings_order_RW2_multi])
```



## Compare ELBO

```{r fig.width=6, fig.height=4}
# they might have different lengths, so we will just plot them all
max_length <- max(length(result_RW2_PC1$elbo), length(result_RW2_pcurve$elbo), length(result_RW2_multi$fits$K_65$elbo))
elbo_PC1 <- c(result_RW2_PC1$elbo, rep(NA, max_length - length(result_RW2_PC1$elbo)))
elbo_PCurve <- c(result_RW2_pcurve$elbo, rep(NA, max_length - length(result_RW2_pcurve$elbo)))
elbo_MultiRes <- c(result_RW2_multi$fits$K_65$elbo, rep(NA, max_length - length(result_RW2_multi$fits$K_65$elbo)))

elbo_values <- data.frame(
  Iteration = 1:max_length,
  ELBO_PC1 = elbo_PC1,
  ELBO_PCurve = elbo_PCurve,
  ELBO_MultiRes = elbo_MultiRes
)

elbo_long <- pivot_longer(elbo_values, cols = -Iteration, names_to = "Method", values_to = "ELBO")

ggplot(elbo_long, aes(x = Iteration, y = ELBO, color = Method, linetype = Method)) +
  geom_line(size = 1) +
  labs(title = "ELBO Convergence Comparison", x = "Iteration", y = "ELBO") +
  theme_minimal()
```






