---
title: "explore_swiss_roll"
author: "Ziang Zhang"
date: "2025-12-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
## Introduction

In this example, we will consider how to learn a 1D manifold (swiss-roll/spiral) embedded in 2D with noise, using various variants of principal curve fitting and our SmoothEM algorithm.

Simulate a 1D swiss-roll (spiral) embedded in 2D with noise.

```{r}
simulate_swiss_roll_1d_2d <- function(n = 800,
                                     t_range = c(1.5 * pi, 6 * pi),
                                     sigma = 0.15,     # can be scalar or length-2
                                     seed = 1) {
  stopifnot(length(t_range) == 2)
  stopifnot(length(sigma) == 1 || length(sigma) == 2)

  set.seed(seed)

  # 1D parameter
  t <- runif(n, min = t_range[1], max = t_range[2])

  # true 2D "swiss roll" (spiral)
  x_true <- t * cos(t)
  y_true <- t * sin(t)

  # noise
  if (length(sigma) == 1) sigma <- rep(sigma, 2)
  x_obs <- x_true + rnorm(n, 0, sigma[1])
  y_obs <- y_true + rnorm(n, 0, sigma[2])

  list(
    t = t,
    truth = data.frame(x = x_true, y = y_true),
    obs   = data.frame(x = x_obs,  y = y_obs)
  )
}

# Example
sim <- simulate_swiss_roll_1d_2d(n = 1500, sigma = 0.2, seed = 123)

# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y", main = "Noisy 2D swiss-roll (spiral)")
# Colors mapped from t
pal <- colorRampPalette(c("navy", "cyan", "yellow", "red"))(256)
t_scaled <- (sim$t - min(sim$t)) / (max(sim$t) - min(sim$t))  # in [0,1]
col_t <- pal[pmax(1, pmin(256, 1 + floor(255 * t_scaled)))]
# Add truth, colored by t
points(sim$truth$x, sim$truth$y, pch = 16, cex = 0.25, col = col_t)
```


## Principal Curve Fitting

The simpliest principal curve is just using the initial PCA line (zero iterations).

```{r}
library(princurve)
# Initial fit with zero iterations (just the initial PCA line)
pc_fit <- principal_curve(as.matrix(sim$obs), maxit = 0)
# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
# Add principal curve (sorted by lambda!)
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
```

Apply principal curve fitting to recover the 1D structure with the default settings.

```{r}
set.seed(123)
pc_fit <- principal_curve(as.matrix(sim$obs))

# Plot noisy observations (grey)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
# Add principal curve (sorted by lambda!)
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
```

Try to increase fitting accuracy by lowering the convergence threshold.

```{r}
set.seed(123)
pc_fit <- principal_curve(as.matrix(sim$obs), thresh = 0.0001, 
                          stretch = 0, maxit = 100)

plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit to Noisy Swiss-Roll",
     asp = 1)
ord <- order(pc_fit$lambda)
lines(pc_fit$s[ord, 1], pc_fit$s[ord, 2], col = "black", lwd = 2)
```


Try a different smoother: lowess.

```{r}
set.seed(123)
pc_fit_lowess <- principal_curve(as.matrix(sim$obs), smoother = "lowess", 
                                 thresh = 0.0001,
                                 maxit = 1000)
plot(sim$obs$x, sim$obs$y, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x", ylab = "y",
     main = "Principal Curve Fit (lowess) to Noisy Swiss-Roll",
     asp = 1)
ord <- order(pc_fit_lowess$lambda)
lines(pc_fit_lowess$s[ord, 1], pc_fit_lowess$s[ord, 2], col = "black", lwd = 2)
```


## SmoothEM

None of the above principal curve fittings are able to recover the 1D swiss-roll structure well, even when we switch to different smoothers.

Most likely it is due to the improper initialization using PCA line.

Now let's try various way to initialize the SmoothEM algorithm to recover the 1D structure.

```{r}
library(tibble)
library(tidyr)
library(ggplot2)
library(Rtsne)
library(umap)
library(Matrix)

set.seed(123)
source("./code/plot_ordering.R")
source("./code/linear_EM.R")
source("./code/general_EM.R")
source("./code/prior_precision.R")
source("./code/kriging.R")

X <- as.matrix(sim$obs)
```

### Initialization using PCA

For comparison, let's also start with PCA initialization.

```{r}
set.seed(1234)
q <- 2
Q_prior_RW2 <- make_random_walk_precision(K = 65, 
                                          d=ncol(X), 
                                          lambda = 30, 
                                          q = q, 
                                          ridge = 0)

PC1 <- prcomp(X,center = TRUE, scale. = FALSE)$x[,1]

init_params <- make_init(X = X, ordering_vec = -PC1, K = 65)
result_RW2_PC1 <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 1000,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

result_RW2_PC1$clustering <- apply(result_RW2_PC1$gamma, 1, which.max)
result_RW2_PC1$position <- result_RW2_PC1$gamma %*% matrix(1:65, ncol=1)
```

```{r}
plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_PC1$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```

Just like what we would expect, the inferred curve is not able to recover the swiss-roll structure, as what we have seen in the principal curve fitting with PCA initialization.

### Initialization using Principal Curve

Now let's take the principal curve scores to initialize the SmoothEM.

```{r}
pc_score <- pc_fit_lowess$lambda
init_params <- make_init(X = X, ordering_vec = -pc_score, K = 65)
result_RW2_pcurve <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 1000,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

result_RW2_pcurve$clustering <- apply(result_RW2_pcurve$gamma, 1, which.max)
result_RW2_pcurve$position <- result_RW2_pcurve$gamma %*% matrix(1:65, ncol=1)
```

```{r}
plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_pcurve$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```

The principal curve initialization also fails to recover the swiss-roll structure.


### Initialization using t-SNE

Let's try t-SNE to get a 1D ordering of the data points for initialization.

```{r}
set.seed(123)
tsne_result <- Rtsne(X, dims = 2, perplexity = 10, verbose = FALSE, max_iter = 500)
init_params <- make_init(X = X, ordering_vec = -tsne_result$Y[,1], K = 65)
result_RW2_tsne <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 1000,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

result_RW2_tsne$clustering <- apply(result_RW2_tsne$gamma, 1, which.max)
result_RW2_tsne$position <- result_RW2_tsne$gamma %*% matrix(1:65, ncol=1)
```

```{r}
plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_tsne$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```


The t-SNE initialization also fails to recover the swiss-roll structure.



### Initialization using Multi-scale SmoothEM

Now, let's try the progressive multi-scale SmoothEM initialization.

```{r fig.width=6, fig.height=4}
result_RW2_multi <- progressive_smoothEM(
  data = X,
  m_max = 6,              
  lambda_final = 100,
  q = 2,
  ridge = 0,
  nugget_kriging = 0,  
  tol = 1e-3,
  max_iter = 1000,
  relative_lambda = T,
  plot_each_stage = F,
  verbose = F
)
result_RW2_multi$clustering <- apply(result_RW2_multi$fits$K_65$gamma, 1, which.max)
result_RW2_multi$position <- result_RW2_multi$fits$K_65$gamma %*% matrix(1:65, ncol=1)
```


```{r}
plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_multi$fits$K_65$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```

Still no luck in recovering the swiss-roll structure.


## Why none of these method work?

It seems like none of these methods provide a good visual recovery of the swiss-roll structure.

One possible reason is that the initialization orderings are all very different from the true underlying ordering (by t), and all of these iterative methods are really easy to get stuck in local optima in the ordering space.

Let's cheat a bit by using the true underlying ordering to initialize the SmoothEM. We should see a much better objective value (ELBO) and visual recovery of the swiss-roll structure.

```{r}
set.seed(1234)
init_params <- make_init(X = X, ordering_vec = -sim$t, K = 65)
result_RW2_cheat <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 1000,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

result_RW2_cheat$clustering <- apply(result_RW2_cheat$gamma, 1, which.max)
result_RW2_cheat$position <- result_RW2_cheat$gamma %*% matrix(1:65, ncol=1)
```

```{r}
plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_cheat$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```

```{r}
range_elbo <- c(
  min(c(result_RW2_PC1$elbo_trace,
        result_RW2_pcurve$elbo_trace,
        result_RW2_tsne$elbo_trace,
        result_RW2_multi$fits$K_65$elbo_trace,
        result_RW2_cheat$elbo_trace)),
  max(c(result_RW2_PC1$elbo_trace,
        result_RW2_pcurve$elbo_trace,
        result_RW2_tsne$elbo_trace,
        result_RW2_multi$fits$K_65$elbo_trace,
        result_RW2_cheat$elbo_trace))
)
# compare ELBOs
plot(NULL, xlim=c(1,300), ylim=range_elbo,
     xlab="Iteration", ylab="ELBO",
     main="ELBO Comparison")
lines(result_RW2_PC1$elbo_trace, col="blue", lwd=2)
lines(result_RW2_pcurve$elbo_trace, col="green", lwd=2)
lines(result_RW2_tsne$elbo_trace, col="purple", lwd=2)
lines(result_RW2_multi$fits$K_65$elbo_trace, col="orange", lwd=2)
lines(result_RW2_cheat$elbo_trace, col="red", lwd=2)
legend("bottomright",
       legend=c("PCA Init", "P-Curve Init", "t-SNE Init", "Multi-scale Init", "Cheat Init"),
       col=c("blue", "green", "purple", "orange", "red"),
       lwd=2)
```


Let's try to see when the SmoothEM fit starts to break down as we alter the initialization ordering to be more and more different from the true ordering.

```{r}
set.seed(1234)
order_vec <- sim$t + rnorm(length(sim$t), sd=3)
# apply some monotone transformation to make it less correlated with true t
f <- function(x) {
  # pnorm(x)
  x
}
order_vec <- f(order_vec)

plot(rank(order_vec), rank(sim$t),
     xlab="Altered Ordering Rank", ylab="True t Rank",
     main="Altered vs True Ordering Ranks")
# show the lm line and its r2 on the plot
abline(lm(rank(sim$t) ~ rank(order_vec)), col="red", lwd=2)
r2 <- summary(lm(rank(sim$t) ~ rank(order_vec)))$r.squared
legend("topleft", legend=paste0("R^2 = ", round(r2, 3)), col="red", lwd=2)

init_params <- make_init(X = X, ordering_vec = order_vec, K = 65)
result_RW2_cheat <- EM_algorithm(
  data = X,
  Q_prior = Q_prior_RW2,
  init_params = init_params,
  max_iter = 1000,
  modelName = "EEI",
  tol = 1e-3,
  eigen_tol = 0,
  rank_deficiency = q * ncol(X),
  nugget = 0,
  relative_lambda = T,
  verbose = F
)

plot(X, cex=1,
     xlab="X", ylab="Y",
     pch=19)
     
# Turn mu_list into matrix
mu_matrix <- do.call(rbind, result_RW2_cheat$params$mu)

# Draw arrows showing sequence of cluster means
for (k in 1:(nrow(mu_matrix)-1)) {
  if (sqrt(sum((mu_matrix[k+1,] - mu_matrix[k,])^2)) > 1e-6) {
    arrows(mu_matrix[k,1], mu_matrix[k,2],
           mu_matrix[k+1,1], mu_matrix[k+1,2],
           col="orange", lwd=4, length=0.1)
  }
}
points(mu_matrix, pch=8, cex=1, lwd=1, col="orange")
```

Now, let's vary the amount of noise added to the true ordering to see how the final ELBO changes with respect to the $R^2$ between the altered ordering and the true ordering.

```{r eval=FALSE}
r2_rank <- c()
elbo_final <- c()
for (sd_noise in seq(0, 5, by=0.1)) {
  set.seed(1234)
  order_vec <- sim$t + rnorm(length(sim$t), sd=sd_noise)
  # apply some monotone transformation to make it less correlated with true t
  f <- function(x) {
    # pnorm(x)
    x
  }
  order_vec <- f(order_vec) 
  r2 <- summary(lm(rank(sim$t) ~ rank(order_vec)))$r.squared
  r2_rank <- c(r2_rank, r2)
  init_params <- make_init(X = X, ordering_vec = order_vec, K = 65)
  result_RW2 <- EM_algorithm(
    data = X,
    Q_prior = Q_prior_RW2,
    init_params = init_params,
    max_iter = 1000,
    modelName = "EEI",
    tol = 1e-3,
    eigen_tol = 0,
    rank_deficiency = q * ncol(X),
    nugget = 0,
    relative_lambda = T,
    verbose = F
  )
  elbo_final <- c(elbo_final, tail(result_RW2$elbo_trace, n=1))
}
save(r2_rank, elbo_final, file="./data/swiss_roll_r2_elbo.RData")
```

Showing the relationship between the final ELBO from SmoothEM and the $R^2$ between the altered initialization ordering and the true ordering.
The amount of noise added to the true ordering increases from 0 to 5 in steps of 0.1, shown as text labels on the plot.

```{r}
load("./data/swiss_roll_r2_elbo.RData")
plot(r2_rank, elbo_final,
     xlab="R^2 between Altered and True Ordering Ranks", ylab="Final ELBO",
     main="Final ELBO vs R^2 of Initialization Ordering")
lines(smooth.spline(r2_rank, elbo_final, df=5), col="red", lwd=2)
# show sd in each point
text(r2_rank, elbo_final, labels=seq(0, 5, by=0.1), pos=3)
```




