---
title: "Exploring how PC works as an ordering initialization"
author: "Ziang Zhang"
date: "2026-09-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                     fig.width = 11, fig.height = 7)
library(MPCurver)
library(ggplot2)
library(fields)
theme_set(theme_minimal(base_size = 12))
```

## Intro

Here we store the simulation configurations, including the number of samples (`n`), number of columns (`p`), and the signal to noise ratio (`snr`).

```{r configuration}
config <- list(
  # Data generation
  seed = 20260932L, n = 300L, p = 12L, snr = 4,
  generator = "matern",       # "matern" or "fourier"

  # For MPCurve fitting
  K = 50L, position_prior = "fixed", # "fixed" = uniform; or "adaptive"
  max_iter = 2000L, tol = 1e-6 
)
```

```{r simulation-module}
simulate_signal <- function(t, p, config) {
  if (config$generator == "matern") {
    covariance <- fields::Matern(abs(outer(t, t, "-")),
                             aRange =  0.1,
                             smoothness = 2.5)
    L <- t(chol(covariance + diag(1e-10, length(t))))
    return(L %*% matrix(rnorm(length(t) * p), length(t), p))
  }
  if (config$generator == "fourier") {
    h <- 2:6; decay <- 1.5;
    angles <- outer(t, pi * h)
    a <- matrix(rnorm(length(h) * p), length(h), p)
    b <- matrix(rnorm(length(h) * p), length(h), p)
    return(sin(angles) %*% (h^(-decay) * a) +
           cos(angles) %*% (h^(-decay) * b))
  }
  stop("Unknown generator: ", config$generator)
}
simulate_data <- function(config) {
  stopifnot(config$n > config$K, config$p >= 1, config$snr > 0)
  set.seed(config$seed)
  t <- runif(config$n)
  signal <- simulate_signal(t, config$p, config)
  stopifnot(identical(dim(signal), c(config$n, config$p)),
            all(is.finite(signal)), all(apply(signal, 2, sd) > 0))
  signal <- scale(signal)  # unit signal variance per column
  # Separate noise seed keeps noise fixed when changing the generator.
  set.seed(config$seed + 1L)
  X <- signal + matrix(rnorm(length(signal)), nrow(signal)) / sqrt(config$snr)
  colnames(X) <- colnames(signal) <- paste0("V", seq_len(config$p))
  list(t = t, signal = signal, X = X)
}
data <- simulate_data(config)
```

```{r plotting-module}
scale01 <- function(x) {
  stopifnot(diff(range(x)) > 0)
  (x - min(x)) / diff(range(x))
}
recovery <- function(x, truth) abs(cor(x, truth, method = "spearman"))

feature_frame <- function(values, position, truth) {
  do.call(rbind, lapply(seq_len(ncol(values)), function(j)
    data.frame(position = position, value = values[, j], truth = truth,
               feature = factor(colnames(values)[j], levels = colnames(values)))))
}

plot_features <- function(data, position, signal_position = NULL,
                          xlabel = deparse(substitute(position))) {
  observed <- feature_frame(data$X, position, data$t)
  if (is.null(signal_position)) {
    ggplot(observed, aes(position, value, color = truth)) +
      geom_point(size = 0.8, alpha = 0.65) +
      scale_color_viridis_c(name = "True t", limits = c(0, 1)) +
      facet_wrap(~feature, ncol = 4) + labs(x = xlabel, y = "Observed value")
  } else {
    idx <- order(data$t)
    paths <- feature_frame(data$signal[idx, , drop = FALSE],
                           signal_position[idx], data$t[idx])
    ggplot(observed, aes(position, value)) +
      geom_point(color = "gray55", size = 0.6, alpha = 0.3) +
      geom_path(data = paths, color = "#0072B2", linewidth = 0.8) +
      facet_wrap(~feature, ncol = 4) + labs(x = xlabel, y = "Feature value")
  }
}
```

```{r figure-1, fig.cap="Figure 1. Features versus true t. Blue paths join noiseless values in true-position order; gray points are observations."}
plot_features(data, data$t, signal_position = data$t,
              xlabel = "True latent position")
```

```{r figure-2, fig.cap="Figure 2. Features versus the reference column. Blue paths are noiseless pairs; gray points contain noise on both axes."}
j <- 1
plot_features(data, data$X[, j], signal_position = data$signal[, j],
              xlabel = sprintf("Observed %s (reference column)", colnames(data$X)[j]))
```

## PCA

We first try using PC1 as the ordering initialization.

```{r pca}
pca <- prcomp(data$X, center = TRUE, scale. = FALSE)
pc_raw <- pca$x[, 1]
direction <- if (cor(pc_raw, data$t, method = "spearman") < 0) -1 else 1
pc_position <- scale01(direction * pc_raw)
```

Take a look at the columns ordered by the PCA positions.

```{r figure-3, fig.cap="Figure 3. Observations versus scaled PC1 scores; colors indicate true t."}
plot_features(data, pc_position, xlabel = "PC1 scores scaled to [0,1]")
```

Compare the PCA positions against the true latent positions.

```{r figure-4, fig.height=5, fig.cap="Figure 4. PCA positions versus true t. Dashed line: identity."}
ggplot(data.frame(truth = data$t, estimate = pc_position), aes(truth, estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray70") +
  geom_point(alpha = 0.65, size = 1.2) + coord_equal() +
  labs(x = "True latent position", y = "Scaled PC1 position",
       title = sprintf("PCA: |Spearman rho| = %.3f", recovery(pc_position, data$t)))
```

## MPCurve

Now let's try using MPCurve to improve the PC initialization.

```{r}
fit <- MPCurver::fit_mpcurve(X = data$X, K = config$K, 
                             position_prior = config$position_prior,
                             method = "PCA",
                             iter = config$max_iter, tol = config$tol)
```

Compute the posterior mean positions from the fitted MPCurve.

```{r fitting-module2}
mp_position <- as.numeric(fit$gamma %*% seq(0, 1, length.out = config$K))
if (direction < 0) mp_position <- 1 - mp_position
initial_elbo <- head(fit$elbo_trace, 1)
final_elbo <- tail(fit$elbo_trace, 1)
# Keep convergence status visible even when the iteration budget is exhausted.
data.frame(version = as.character(packageVersion("MPCurver")),
           iterations = fit$iter, converged = fit$converged,
           initial_ELBO = initial_elbo, final_ELBO = final_elbo)
```

Plot the columns ordered by the MPCurve posterior mean positions.

```{r figure-5, fig.cap="Figure 5. Observations versus MPCurve posterior mean positions; colors indicate true t."}
plot_features(data, mp_position, xlabel = "MPCurve posterior mean position")
```

Compare the MPCurve posterior mean positions against the true latent positions.

```{r figure-6, fig.height=5.8, fig.cap="Figure 6. Before/after comparison against true t, with the same orientation. Initial ELBO is evaluated after PCA quantile discretization and variational initialization, before CAVI updates."}
labels <- c(
  sprintf("PCA initialization\nrho = %.3f | ELBO = %.2f",
          recovery(pc_position, data$t), initial_elbo),
  sprintf("MPCurve: %d updates, converged = %s\nrho = %.3f | ELBO = %.2f",
          fit$iter, fit$converged, recovery(mp_position, data$t), final_elbo))
comparison <- data.frame(
  truth = rep(data$t, 2), estimate = c(pc_position, mp_position),
  stage = factor(rep(labels, each = length(data$t)), levels = labels))
ggplot(comparison, aes(truth, estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray70") +
  geom_point(alpha = 0.65, size = 1.2) + facet_wrap(~stage, nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "True latent position", y = "Estimated latent position")
```


## PCA across P

Let's change P from 12 to 200, each with 30 replications, and see how the performance of PCA changes.

```{r pca-vs-p, fig.height=5, fig.cap="Figure 7. |Spearman rho| between PC1 and true t across P; each point is one replication."}
# P = 12, 30, 50, 80, 100, 200
# show boxplot of spearman correlation between PCA and true t in each replication, across different P values
p_values <- c(12L, 30L, 50L, 80L, 100L, 200L)
n_rep <- 30L
pca_study <- do.call(rbind, lapply(p_values, function(p) {
  do.call(rbind, lapply(seq_len(n_rep), function(r) {
    # Step seeds by 2: simulate_data() uses seed and seed + 1 (noise).
    cfg <- modifyList(config, list(p = p, seed = config$seed + 2L * r))
    sim <- simulate_data(cfg)
    pc1 <- prcomp(sim$X, center = TRUE, scale. = FALSE)$x[, 1]
    data.frame(p = p, replicate = r, rho = recovery(pc1, sim$t))
  }))
}))
aggregate(rho ~ p, pca_study, function(x) c(mean = mean(x), sd = sd(x)))
ggplot(pca_study, aes(factor(p), rho)) +
  geom_boxplot(outlier.shape = NA, fill = "gray90") +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.2) +
  labs(x = "Number of features P", y = "|Spearman rho| (PC1 vs true t)",
       title = sprintf("PCA recovery across P (%d replications each)", n_rep))
```
