---
title: "Analysis: Pancreatic Cell Loadings"
author: "Ziang Zhang"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = paste0(
  tools::file_path_sans_ext(basename(knitr::current_input())),
  "_files/figure-html/"
))
result_dir <- "experiments/mpcurve_v030_site/results/pancreas"
m1 <- readRDS(file.path(result_dir, "fit_m1.rds"))
m2 <- readRDS(file.path(result_dir, "fit_m2.rds"))
samples <- read.csv(file.path(result_dir, "sample_positions.csv"), check.names = FALSE)
assignments <- read.csv(file.path(result_dir, "factor_assignments.csv"))
X <- m1$fit$data
factor_columns <- c(3, 8, 9, 12, 17, 18, 20, 21)
stopifnot(identical(dim(X), c(865L, 8L)),
          isTRUE(all.equal(X, m2$fit$data, check.attributes = FALSE)))
position_correlations <- c(
  A = abs(cor(samples$M1, samples$M2_A, method = "spearman")),
  B = abs(cor(samples$M1, samples$M2_B, method = "spearman"))
)
```

## Data and analysis question

We analyze semi-NMF loading factors from pancreatic cells. We sample at most
500 cells per cell type with seed 1, retain cells measured with `inDrop3`,
and select factor columns 3, 8, 9, 12, 17, 18, 20, and 21. This gives
**865 cells by 8 factors**. We fit $M=1$ and $M=2$ MPCurve models to the same
matrix to see what each specified number of orderings captures.

## Fitting settings

Both fits use CAVI, an RW2 trajectory prior, $K=50$ grid positions, and
`ridge = 0`. The $M=1$ fit starts from PCA; the $M=2$ fit uses Fiedler
ordering and similarity-based feature initialization. Exact seeds and
iteration budgets are in the [reproduction script](https://github.com/AgueroZZ/InferOrder/blob/master/experiments/mpcurve_v030_site/run_pancreas.R).

```{r fit-settings-table, echo=FALSE}
settings <- read.csv(file.path(result_dir, "fit_settings.csv"))
knitr::kable(settings[, c("model", "seed", "method", "K", "rw_q", "ridge")],
             caption = "Table 1. Fixed model and initialization settings.")
```

## One versus two orderings

The plots below arrange the same loadings according to each fitted sample
ordering. The first panel uses $M=1$; the other two are orderings A and B
from the fixed $M=2$ fit. Colours show the observed loading values, with the
same colour scale in all panels. Ordering direction is arbitrary.

```{r ordered-loadings, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Figure 1. Eight semi-NMF loading factors arranged by the inferred cell positions under fixed M = 1 and M = 2 fits."}
plot_loading_heatmap <- function(position, title) {
  ordered_X <- X[order(position), , drop = FALSE]
  colours <- grDevices::colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(100)
  image(seq_len(nrow(ordered_X)), seq_len(ncol(ordered_X)), ordered_X,
        col = colours, zlim = range(X), yaxt = "n",
        xlab = "Cells ordered by inferred pseudotime",
        ylab = "Semi-NMF factor column", main = title)
  axis(2, at = seq_len(ncol(X)), labels = factor_columns, las = 2)
  box()
}
old_par <- par(mfrow = c(3, 1), mar = c(4, 6, 3, 1))
plot_loading_heatmap(samples$M1, "M = 1")
plot_loading_heatmap(samples$M2_A, "M = 2, ordering A")
plot_loading_heatmap(samples$M2_B, "M = 2, ordering B")
par(old_par)
```

The absolute Spearman correlations between the $M=1$ positions and the
$M=2$ positions are **`r sprintf("%.3f", position_correlations["A"])`** for
ordering A and **`r sprintf("%.3f", position_correlations["B"])`** for
ordering B. Ordering A closely follows the single-ordering fit; ordering B
places cells differently.

Cell-type annotations provide another view of the fitted positions. Each
point in Figure 2 is one cell, grouped by its recorded cell type.

```{r cell-types, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Figure 2. Inferred pseudotime for each pancreatic cell, grouped by the recorded cell type."}
type_levels <- names(sort(table(samples$cell_type), decreasing = TRUE))
plot_cell_types <- function(position, title) {
  y <- match(samples$cell_type, type_levels)
  plot(position, jitter(y, amount = 0.22), pch = 16, cex = 0.55,
       col = grDevices::adjustcolor("#2E8A99", alpha.f = 0.38),
       xlab = "Inferred pseudotime", ylab = "Cell type", yaxt = "n",
       ylim = c(length(type_levels) + 0.6, 0.4), main = title)
  axis(2, at = seq_along(type_levels), labels = type_levels,
       las = 2, cex.axis = 0.65)
  grid(nx = NA, ny = length(type_levels), col = "grey90")
}
old_par <- par(mfrow = c(3, 1), mar = c(4, 9, 3, 1))
plot_cell_types(samples$M1, "M = 1")
plot_cell_types(samples$M2_A, "M = 2, ordering A")
plot_cell_types(samples$M2_B, "M = 2, ordering B")
par(old_par)
```

## Feature assignments in the two-ordering fit

```{r assignment-table, echo=FALSE}
knitr::kable(assignments, digits = 3,
             caption = "Table 2. Posterior ordering probabilities for each selected semi-NMF factor.")
```

Seven factors follow ordering A; only factor 20 follows B. Most cells have
similar fitted positions on B, while a smaller subset spreads across its
range. The second ordering is therefore driven mainly by factor 20 in this
fit.

## Fitting objectives

```{r objective-traces, echo=FALSE, fig.width=10, fig.height=4.3, fig.cap="Figure 3. Within-model objectives: the M = 1 CAVI ELBO and the M = 2 structural objective at T = 1."}
t1_objective <- m2$fit$objective_history[m2$fit$temperature_history == 1]
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(m1$fit$elbo_trace, type = "l", lwd = 2, col = "#2E8A99",
     xlab = "Recorded iteration", ylab = "ELBO", main = "M = 1")
plot(seq_along(t1_objective), t1_objective, type = "l", lwd = 2,
     col = "#E76F51", xlab = "Recorded T = 1 update",
     ylab = "Structural objective", main = "M = 2")
par(old_par)
```

Each objective rises during its own fit. The two panels use different
objectives, so their heights cannot be used to choose $M$.

## Reproduce

The [reproduction
script](https://github.com/AgueroZZ/InferOrder/blob/master/experiments/mpcurve_v030_site/run_pancreas.R)
reconstructs the input matrix and both fits. The [saved
results](https://github.com/AgueroZZ/InferOrder/tree/master/experiments/mpcurve_v030_site/results/pancreas)
include selected cell IDs, fit settings, assignments, and fitted positions.
