---
title: "Simulation: One Latent Ordering"
author: "Ziang Zhang"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = paste0(
  tools::file_path_sans_ext(basename(knitr::current_input())),
  "_files/figure-html/"
))
result_path <- "experiments/mpcurve_v030_site/results/simulations/simulation_m1.rds"
result <- readRDS(result_path)
stopifnot(inherits(result$fit, "mpcurve"), identical(result$fit$algorithm, "cavi"))
sim <- result$simulation
fit <- result$fit
X <- as.matrix(sim$obs)
position <- result$position
summary <- result$summary
```

## Question and design

Can a single MPCurve ordering recover a known sample trajectory from noisy
two-dimensional observations? We use the spiral generator from the
[MPCurver getting-started
article](https://aguerozz.github.io/MPCurver/articles/mpcurve_intro.html):
1,000 samples with latent $t\in[1.5\pi,4.5\pi]$ and Gaussian noise standard
deviation 0.12. The simulation seed is 1. A CAVI fit uses Fiedler
initialization, $K=60$ grid positions, and at most 120 iterations.

## Ordering recovery

```{r recovery, echo=FALSE, fig.width=10, fig.height=4.5, fig.cap="Figure 1. The simulated spiral coloured by the generating position (left) and the inferred posterior-mean grid position (right)."}
palette <- grDevices::colorRampPalette(
  c("#1F3A5F", "#2E8A99", "#F2C14E", "#E76F51")
)(256)
position_colours <- function(values) palette[cut(values, breaks = 256, labels = FALSE)]
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(X, col = position_colours(sim$t), pch = 19, cex = 0.55,
     xlab = "Feature 1", ylab = "Feature 2", main = "Generating position")
plot(X, col = position_colours(position), pch = 19, cex = 0.55,
     xlab = "Feature 1", ylab = "Feature 2", main = "Inferred position")
par(old_par)
```

The absolute Spearman correlation between true $t$ and inferred
posterior-mean grid position is **`r sprintf("%.4f", summary$abs_spearman)`**. The
absolute value accounts for the arbitrary direction of an inferred ordering.
This result comes from one simulated data set.

## Objective trace

```{r elbo, echo=FALSE, fig.width=7, fig.height=4.2, fig.cap="Figure 2. Evidence lower bound recorded during the single-ordering CAVI fit."}
plot(fit$elbo_trace, type = "l", lwd = 2, col = "#2E8A99",
     xlab = "Recorded iteration", ylab = "ELBO")
```

The ELBO increases through the fit and levels off as CAVI converges.

## Reproduce

The [experiment
script](https://github.com/AgueroZZ/InferOrder/blob/master/experiments/mpcurve_v030_site/run_simulations.R)
regenerates the data, fit, and figures with MPCurver 0.3.0. The
[saved results](https://github.com/AgueroZZ/InferOrder/tree/master/experiments/mpcurve_v030_site/results/simulations)
include the generating truth and fitted model.
